# (PART) Modules {-}

# Practical 1

## Requirements
- R and RStudio must be installed prior to the session - you can follow the instructions below
  - [Install R](https://cran.r-project.org/)
  - [Install RStudio](https://posit.co/download/rstudio-desktop/)
  
## Lecture
<iframe 
  src="https://docs.google.com/presentation/d/10LFAUpZFSsJ2ZsJH6y70J7RfVGb97OfR/preview" 
  width="640" 
  height="480" 
  frameborder="0" 
  allow="autoplay">
</iframe>

<!--<iframe width="640" height="360" src="YOUTUBE EMBED LINK" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen></iframe>
-->

## Lab

<!--Materials for this practical can be downloaded [from here](https://github.com/MJousse/IntroToR/).-->

### 1. R basics
In the first module of the workshop, the goals are to familiarize with the language and the logic behind it; Get started with R studio and create your first project; Create your first `.R` file to write down the live code; and get comfortabble with installing packages.

**Module content:**
-   Seeking help
-   Installing packages

[**Script**](./exercises/exercises_module1/scripts/01GettingStarted.Rmd)

#### R Markdown File

R Markdown files act as notebooks which can make it easier to document and share code. All the results from the code as well as your comments and descriptions can be combined into one, tidy document. This makes it easy to combine insights, protocols and figures in a practical and organized fashion.

#### Setting up:

##### Installing a package

To install a package, we simply use the function `install.packages()` with the name of the desired package in quotation marks (i.e. as a string!)

```{r, eval=F}
install.packages("packageName1")

# for multiple packages:
install.packages(c("packageName1", "packageName2"))

```

##### Loading packages

To load a package you simply use the `library()` function. Note function takes the name of an installed library without quotation marks.

```{r, eval=F}

library(packageName1)

```

#### R Basics:

##### Variable names

There are several naming conventions in coding: `snake_case` and `CamelCase` are the most common. There is no particular preference in R for naming, but make sure you are **consistent** and that your variable names are **meaningful**.

##### Case sensitivity

R is case sensitive. This means that `var1` and `Var1` represent two different objects.

##### 1 based indexing

Indexing (calling elements in a list) in R starts with 1. The first element has index 1, the second 2, etc..

```{r, eval=F}
# Initialising a vector
var1 <- c('a', 'b', "c", "d") # note ' and " are the same in R :)

# Here we call elements 1 and 2 of var1
var1[1]
var1[2]

#Now we look at how long var1 is using the length() function

length(var1)

```

#### Exercises

##### 1. Create a file

Create a new .R script in R studio. Save it as `Script01` in the `IntroToR` folder.

##### 2. Install base packages

In `Script01`, install the following packages: `"data.table"`,`"datasets"`,`"devtools"`,`"dplyr"`,`"ggplot2"`,`"plyr"`,`"medicaldata"`,`"gapminder"`,`"RColorBrewer"`,`"rmarkdown"`,`"stringr"`,`"tidyr"`,`"tidyverse"`,`"viridis"`.

##### 3. Load a package

Load all packages.

#### Solutions

##### 1. Creating a file

##### 2. Install base packages

```{r, eval = F}
install.packages(c("data.table","datasets","devtools","dplyr","ggplot2","plyr","medicaldata","gapminder","RColorBrewer","rmarkdown","stringr","tidyr","tidyverse","viridis"))
```

##### 3. load packages

```{r, eval = F}
library(data.table)
library(datasets)
library(devtools)
library(dplyr)
library(ggplot2)
library(plyr)
library(medicaldata)
library(gapminder)
library(RColorBrewer)
library(rmarkdown)
library(stringr)
library(tidyr)
library(tidyverse)
library(viridis)
```

### 2. Data types: attributes and built-in functions

In this section participants will understand the differences between classes, objects and data types in R; Compute arithmetic operations; Use logical operators; create objects of different types, learn about their attributes and apply some built-in functions in R; Subset and index objects; and get comfortable with vectorized operations.

**Module content:**
-   Vectors
-   Lists
-   Factors
-   Data frames
-   Arrays
-   Coercion

[**Script**](./exercises/exercises_module1/scripts/02DataTypes.Rmd)


#### Examples
In this section we go over the material.

##### **Comments**

Comments are embedded into your code to help explain its purpose. This is extremely important for reproducibility and for documentation. To start a comment line, you use : `#`. The whole line is commented. Multi-line comments are available using `#'`.

```{r, eval = F}
# This is a comment line 
# print('hello world')

#' start comment
#' 
#' end comment!

```

##### Creating variables

You 'assign' a value to a variable using the assigning operator: `<-`

```{r, eval = F}
# The convention is to use left hand assignation 
var1 <- 12
var2 <- "hello world"
```

Hint: Look at your environment!

##### Printing Variables

```{r, eval = F}
var1
var2

print(var1)
print(var2)
```

```{r, eval = F}
# It is also possible to use the '=' sign, but is NOT convention and therefore not good practice.
var1 = 13
var2 = "hello world"
var1
var2
```

```{r, eval = F}
# Now what happens when you use the function rm()? hint: look at your environment!
rm(var1)
```

##### Data types and data structures

###### Atomic Classes

Atomic classes are the fundamental data type found in R. All subsequent data structures are used to store entries of different atomic classes.

####### 1.Numeric

They store numbers as `double`, and it is stored with decimals. The term double refers to the number of bytes required to store it. Each double is accurate up to 16 significant digits.

####### 2.Integer

They store numbers that can be written without a decimal component. Adding an *L* after an integer tells R to store it as an integer class instead of a numeric

####### 3.Logical

They store the outputs of logical statements - TRUE or FALSE. Can be converted to integer where TRUE = 1 and FALSE = 0.

####### 4.Character

Represents text. Can either be a single character or a word/sentence.

####### 5.Missing Value

Used by R to indicate a missing data entry. Useful for manipulating data sets where missing entries are common.

###### Arithmetic Operations

```{r, eval = F}
# Addition  
2+100000
```

```{r, eval = F}
# Subtraction  
3-5
```

```{r, eval = F}
# Multiplication  
71*9
```

```{r, eval = F}
# Division  
90/((3*5) + 4)
```

```{r, eval = F}
# Power  
2^3
```

###### Logical operators

```{r, eval = F}
# First create two numeric variables  
var1 <- 35  
var2 <- 27
```

```{r, eval = F}
# Equal to  
var1 == var2
```

```{r, eval = F}
# Less than or equal to  
var1 <= var2   

# Not equal
var1 != var2
```

```{r, eval = F}
# They also work with other classes  
var1 <- "mango" 
var2 <- "mangos"
```

```{r, eval = F}
var1 == var2
```

Strings are compared character by character until they are not equal or there are no more characters left to compare.

```{r, eval = F}
var1 < var2
```

We can test if a variable is contained in another object using the contained operator `%in%`.

```{r, eval = F}
"c" %in% letters  
"c" %in% LETTERS

# note letters and LETTERS are built in constants. letters is the lowercase alphabet as a vector, and LETTERS uppercase.
```

##### Data Structures

###### Vectors

Key points:\
- Can only contain objects of the same class\
- Most basic type of R object\
- Variables are vectors

####### 1.Numeric

Creating a numeric vector using `c()`

```{r, eval = F}
x <- c(0.3, 0.1)
x
```

Using the `vector()` function

```{r, eval = F}
x <- vector(mode = "numeric",length = 10)
x
```

Using the `numeric()` function

```{r, eval = F}
x <- numeric(length = 10)
x
```

Creating a numeric vector with a sequence of numbers

```{r, eval = F}
# x <- seq(1,10,1)
# x

x <- seq(1,10,2)
x

x <- rep(2,10)
x
```

Check length of vector with `length()`

```{r, eval = F}
x
length(x)

y <- rep(2,5)
y
length(y)

length(x) == length(y)

```

####### 2.Integer

Creating an integer vector using `c()`

```{r, eval = F}
x <- c(1L,2L,3L,4L,5L)  
x
```

Creating an integer vector of a sequences of numbers

```{r, eval = F}
x <- 1:10
x
```

####### 3.Logical

Creating a logical vector with `c()`

```{r, eval = F}
x <- c(TRUE,FALSE,T,F)
x
```

Creating a logical vector with `vector()`

```{r, eval = F}
x <- vector(mode = "logical",length = 5)
x
```

Creating a logical vector using `logical()`

```{r, eval = F}
x <- logical(length = 10)
x
```

####### 4.Character

```{r, eval = F}
x<-c("a","b","c")
x

x<-vector(mode = "character",length=10)
x

x<-character(length = 3)
x
```

Some useful functions to modify strings

```{r, eval = F}
tolower(LETTERS)
toupper(letters)
paste(letters,1:length(letters),sep="_") # Note the implicit coercion
```

####### 5.Vector attributes

The elements of a vector can have names

```{r, eval = F}
x<-1:5
names(x)<-c("one","two","three","four","five")
x

x<-logical(length = 4)
names(x)<-c("F1","F2","F3","F4")
x
```


###### Built-in functions

**To inspect the contents of a vector**

```{r, eval = F}
is.vector(x) # Check if it is a vector
is.na(x) # Check if it is empty
is.null(x) # Check if it is NULL
is.numeric(x) # Check if it is numeric
is.logical(x) # Check if it is logical
is.character(x) # Check if it is character
```

**To know what kind of vector you are working with**

```{r, eval = F}
class(x) # Atomic class type
typeof(x) # Object type or data structure (matrix, list, array...)
str(x)
```

**To know more about the data contained in the vector**

###### Mathematical operations

```{r, eval = F}
sum(x)
```

```{r, eval = F}
min(x) 
max(x)
```

```{r, eval = F}
x <- seq(1,10,1)
mean(x) 
median(x) 
sd(x)
```

```{r, eval = F}
log(x) 
exp(x)
```

Other operations

```{r, eval = F}
length(x)
table(x)
summary(x)
```

Grouping elements in a vector using `tapply`

```{r, eval = F}
measurements<-sample(1:1000,6) 
samples<-factor(c(rep("case",3),rep("control",3)), 
                levels = c("control", "case"))
```

```{r, eval = F}
tapply(measurements, samples, mean)
```

####### Vector Operations

```{r, eval = F}
x<-1:10
y<-11:20
```

```{r, eval = F}
x*2
```

```{r, eval = F}
x+y

```

```{r, eval = F}
x*y

```

```{r, eval = F}
x^y
```

####### Recycling

If one of the vectors is smaller than the other, operations are still possible. R will replicate the smaller vector to enable the operation to occur.

**IMPORTANT: if the larger vector is NOT a multiple of the smaller vector, the replication will still occur but will end at the length of the larger vector.**

```{r, eval = F}
x<-1:10
y<-c(1,2,3)
x+y
```

###### Indexing and subsetting

For this example, lets create a vector of random numbers from 1 to 100 of size 15.

```{r, eval = F}
x<-sample(x = 1:100,size = 15,replace = F) 
x
```

Using the index/position

```{r, eval = F}
x[1] # Get the first element
x[13] # Get the thirteenth element
```

Using a vector of indices

```{r, eval = F}
x[1:12] # The first 12 numbers
x[c(1,5,6,8,9,13)] # Specific positions only

names(x) <- letters[1:length(x)]

x[c('a','c','d')]
```

Using a logical vector

```{r, eval = F}
# Only numbers that are less than or equal to 10
x<10
x[x>95] 
# 
# # Only even numbers 
# x%%2 == 0
# x[x%%2 == 0]
```

```{r, eval = F}
x<10
x[x<=10] # Only numbers that are less than or equal to 10
```

Skipping elements using indices

```{r, eval = F}
x[c(-1, -5)]
```

Skipping elements using names

```{r, eval = F}
x<-1:10
names(x)<-letters[1:10]
x[names(x) != "a"]
```


###### Factors

Key points:

-   Useful for categorical data
-   Can have implicit order, if needed
-   Each element has a label or level
-   Some operations behave differently on factors

Creating factors with `factor`

```{r, eval = F}
cols<-factor(x = c(rep("red",4),
                   rep("blue",5),
                   rep("green",2)),              
             levels = c("red","blue","green"))
cols
```

```{r, eval = F}
samples <- c("case", "control", "control", "case") 
samples 
samples_factor <- factor(samples, levels = c("control", "case")) 
samples_factor 
str(samples_factor)
```


##### Lists

Key points:\
- Can contain objects of multiple classes\
- Extremely powerful when combined with some R built-in functions

Creating lists with different data types

```{r, eval = F}
l <- list(1:10, list("hello",'hi'), TRUE)
l
```

Assigning names as we create the list

```{r, eval = F}
l<-list(title = "Numbers", 
        numbers = 1:10, 
        logic = TRUE )
l
names(l)

l$numbers
```

###### Indexing and subsetting

Using `[[]]` instead of `[]`

```{r, eval = F}
l[[1]]
```

Using `$` for named lists

```{r, eval = F}
l$logic
```

###### Built-in functions

```{r, eval = F}
l<-list(sample(1:100,10),
        sample(1:100,10),
        sample(1:100,10))
names(l)<-c("r1","r2","r3")
l
```

Performing operations on all elements of the list using `lapply`

```{r, eval = F}
lsums<-lapply(l,sum)
lsums

lsums <- lapply(l,function(a){
  sum(a)^2
})
lsums
```

##### Matrices

Creating a matrix full of zeros with `matrix()`

```{r, eval = F}
m<-matrix(0, ncol=6, nrow=3)
m
class(m)
typeof(m)
```

Creating a matrix from a vector of numbers

```{r, eval = F}
m<-matrix(1:5, ncol=2, nrow=5)
m
```

###### Attributes

Names of each dimension

```{r, eval = F}
colnames(m)<-letters[1:2]
rownames(m)<-LETTERS[1:5]
m
str(m)
```

###### Built-in functions

To know the size of the matrix

```{r, eval = F}
dim(m)
ncol(m)
nrow(m)
```

##### Data frames

Key points:

-   Columns in data frames are vectors

-   Each column can be of a different data type

-   A data frame is essentially a list of vectors

Creating a data frame using `data.frame()`

```{r, eval = F}
df<-data.frame(numbers=1:10,
               low_letters=letters[1:10],
               logical_values=rep(c(T,F),each=5))
df
class(df)
typeof(df)
str(df)
```

Re-naming columns

```{r, eval = F}
colnames(df)[2]<-"lowercase"
head(df)
View(df)
```

###### Indexing and sub-setting

```{r, eval = F}
df$numbers
df["numbers"]
df[1,]
df[,1]

df[3,3]
```

##### Coercion

Converting between data types with `as.` functions

```{r, eval = F}
x<-1:10
as.list(x)
```

```{r, eval = F}
l<-list(numbers=1:10,
        lowercase=letters[1:10])
l
typeof(l)
df<-as.data.frame(l)
df
typeof(df)
```

#### Exercises
Either complete these in the `.Rmd` file or create a new `.R` script and complete them.

##### 1. Write a piece of code that stores a number in a variable and then check if it is greater than 5. Try to use comments!

Bonus: Is there a way to store the result of checking if the number is greater than 5?

```{r, eval = F}
```

##### 2. Make a vector with the numbers 1 through 26. Multiply the vector by 2, and give the resulting vector names A through Z

hint: there is a built in vector called LETTERS

```{r, eval = F}

```

##### 3. Make a matrix with the numbers 1:50, with 5 columns and 10 rows. Did the matrix function fill your matrix by column, or by row, as its default behavior? Once you have figured it out, try to change the default.

hint: read the documentation for `matrix`. To automatically call in documentation type `?` before the function, package, etc.

```{r, eval = F}

```

####### Bonus: Which of the following commands was used to generate the matrix below?

-   `matrix(c(4, 1, 9, 5, 10, 7), nrow = 3)`
-   `matrix(c(4, 9, 10, 1, 5, 7), ncol = 2, byrow = TRUE)`
-   `matrix(c(4, 9, 10, 1, 5, 7), nrow = 2)`
-   `matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)`

```{r, eval = F}

```

##### 4. Create a list of length two containing a character vector for each of the data sections: (1) Data types and (2) Data structures. Populate each character vector with the names of the data types and data structures, respectively.

```{r, eval = F}

```

##### 5. Data frames There are several subtly different ways to call variables, observations and elements from data frames. Try them all and discuss with your team what they return.

Hint, use the function `typeof()`

-   `iris[1]`
-   `iris[[1]]`
-   `iris$Species`
-   `iris["Species"]`
-   `iris[1,1]`
-   `iris[,1]`
-   `iris[1,]`

```{r, eval = F}

```

##### 6. Take the list you created in 4 and coerce it into a data frame. Then change the names of the columns to "dataTypes" and "dataStructures"

```{r, eval = F}

```

##### 7. Create a vector `x` of numbers from 1 to 100. Find all the odd numbers in `x`

Hint: `3 %% 2 = 1` and `4 %% 2 = 0`

```{r, eval = F}


```


#### Solutions

##### 1. Write a piece of code that stores a number in a variable and then check if it is greater than 5. Try to use comments!

Bonus: Is there a way to store the result after checking the number?

```{r, eval = F}

x <- 10
x > 5

#Bonus
y <- x > 5

print(y)
```

##### 2. Make a vector with the numbers 1 through 26. Multiply the vector by 2, and give the resulting vector names A through Z

hint: there is a built in vector called LETTERS

```{r, eval = F}
x <- 1:26
x <- x * 2
names(x) <- LETTERS
```

##### 3. Make a matrix with the numbers 1:50, with 5 columns and 10 rows. Did the matrix function fill your matrix by column, or by row, as its default behavior? Once you have figured it out, try to change the default.

hint: read the documentation for `matrix`. To automatically call in documentation type `?` before the function, package, etc.

```{r, eval = F}
# By default the matrix is filled by columns, we can change this behavior using byrow=TRUE
m<-matrix(1:50,ncol = 5,nrow = 10,byrow = T)
```

####### Bonus: Which of the following commands was used to generate the matrix below?

-   `matrix(c(4, 1, 9, 5, 10, 7), nrow = 3)`
-   `matrix(c(4, 9, 10, 1, 5, 7), ncol = 2, byrow = TRUE)`
-   `matrix(c(4, 9, 10, 1, 5, 7), nrow = 2)`
-   `matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)`

```{r, eval = F}
matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)
```

##### 4. Create a list of length two containing a character vector for each of the data sections: (1) Data types and (2) Data structures. Populate each character vector with the names of the data types and data structures, respectively.

```{r, eval = F}
dt <- c('double', 'complex', 'integer', 'character', 'logical')
ds <- c('data.frame', 'vector', 'factor', 'list', 'matrix')
data.sections <- list(dt, ds)
```

##### 5. Data frames There are several subtly different ways to call variables, observations and elements from data frames. Try them all and discuss with your team what they return.

Hint, use the function `typeof()`

-   `iris[1]`
-   `iris[[1]]`
-   `iris$Species`
-   `iris["Species"]`
-   `iris[1,1]`
-   `iris[,1]`
-   `iris[1,]`

```{r,eval=F}
# The single brace [1] returns the first slice of the list, as another list. In this case it is the first column of the data frame.
iris[1]
# The double brace [[1]] returns the contents of the list item. In this case it is the contents of the first column, a vector of type factor.
iris[[1]]
# This example uses the $ character to address items by name. Species is a vector of type factor.
iris$Species
# A single brace ["Species"] instead of the index number with the column name will also return a list like in the first example
iris["Species"]
# First element of first row and first column. The returned element is an integer
iris[1,1]
# First column. Returns a vector
iris[,1]
# First row. Returns a list with all the values in the first row.
iris[1,]
```

##### 6. Take the list you created in 4 and coerce it into a data frame. Then change the names of the columns to "dataTypes" and "dataStructures"

```{r, eval = F}
df<-as.data.frame(data.sections)
colnames(df)<-c("dataTypes","dataStructures")
```


##### 7. Create a vector `x` of numbers from 1 to 100. Find all the odd numbers in `x`

Hint: `3 %% 2 = 1` and `4 %% 2 = 0`

```{r, eval = F}

x <- c(1:100)
x[x %% 2 ==1]

```



### 3. Control structures
In this module, participants learn to use if-else, while, and for loops; as well as building their own functions; and what packages are.

**Module Content**
* If..Else
* While loop
* For loop
* Functions
* Packages

[**Script**](./exercises/exercises_module1/scripts/03ControlStructuresAndFunctions.Rmd)

#### Examples
In this section we go over the material.

##### If... Else
We can tell our code to perform certain tasks if a certain condition is met.

```{r, eval = F}
x <- 5

#If the number is greater than 5, get its square
if (x > 5){
  x^2
} else{
  #If not, multiply it by 2
  x*2
}
```

```{r, eval = F}
# Can incorporate multiple conditions
x <- 7

if (x > 5 && x < 10){
  #If x is greater than 5 AND is less than 10
  x^2
} else if ( x < 0 || x > 10){
  #If x is less than 0 OR greater than 10
  -x
} else {
  x * 2
}
```

##### For & While Loops

For loops let you iterate over the elements of a vector

```{r, eval = F}
#Iterate from 1 to 10

for (i in 1:10){
  print(i)
}
```

```{r, eval = F}
#Iterate through strings
x <- c('a','ab','cab','taxi')

for (i in x){
  print(i)
}
```

While loops let you iterate over a piece of code until a condition is no longer met.

```{r, eval = F}
x <- 5
#while x is greater than 0
while (x > 0){
  print(x)
  x <- x -1
}
```

##### Functions

Sometimes, we need to repeatedly use a sequence of code. Creating functions helps reduce the amount of clutter in our code.

```{r, eval = F}
#Example Function that calculates mean

new_mean <- function(values){
  print(values)
  #return()
  sum(values)/length(values)
}

x <- 1:5

mean(x) == new_mean(x)

```

##### Installing packages

There are multiple sources and ways to do this.

###### CRAN

```{r,eval=FALSE}
install.packages(c("dplyr","ggplot2","gapminder","medicaldata"))
```

###### BioConductor

For more details about the project you can visit https://www.bioconductor.org

To install packages from BioConductor you first need to install BioConductor itself.

``` {r,eval=F}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(version = "3.16")
```

Then you can install any package you want by using the install

```{r,eval=F}
BiocManager::install("DESeq2")
```

###### GitHub

If you want to install the development version of a package, or you are installing something that is only available on GitHub you can use devtools

```{r,eval=F}
devtools::install_github('andreacirilloac/updateR')
```


#### Exercises

##### 1. Initialise variable `x` with numbers 1 through 10. Calculate the sum of x using base R. Then write a function to perform the sum of `x`.

```{r, eval = F}


```

##### 2. Create a function that returns `TRUE` if the number is even, and `FALSE` if the number passed is odd. Test it.
Hint: Use `if` and `else`!

```{r, eval = F}


```

#### Solutions
Note these solutions are not unique :)

##### 1. Initialise variable `x = c(1,2,3,4,5,6,7,8,9,10)` using base R. Calculate the sum of x. Then write a function to perform the sum of `x`.

```{r, eval = F}

x <- c(1,2,3,4,5,6,7,8,9,10)

sum(x)


new.sum <- function(x){
  
  len <- length(x) # also base R!
  
  y <- 0
  
  for(i in 1:len){
    
    y <- y + x[i]

  }
  
  return(y)
  
}

new.sum(x)

```


##### 2. Create a function that returns `TRUE` if the number is even, and `FALSE` if the number passed is odd. Test it.
Hint: Use `if` and `else`!

```{r, eval = F}

is.even <- function(x){
  
  if(x %%2 == 0){
    y <- TRUE
  }else{
    y <- FALSE
  }
  
  return(y)
}


```


### 4. Basic data manipulation

In this module participants will learn how to read/write data to/from files with different formats (.tsv, .csv); become familiar with basic data-frame operations; (3) index and subset data frames using base R; manipulate individual data frame columns.

**Module content:**
-   Reading/writing data
-   Exploring data frames

[**Script**](./exercises/exercises_module1/scripts/04BasicDataManipulation.Rmd)

#### Examples

In this section we go over the material.

##### `.Rmd` files and paths

Despite being in a `.Rproj`, the working directory for a `.Rmd` is automatically the folder the `.Rmd` file is in.

```{r, eval = F}
getwd() #function to get the working directory
```

Therefore, to save a file into a desired path becomes a bit more complicated. Fortunately the following shorthands exist: `.` signifies the working directory, therefore setting a path as `./` will save the object to the current directory. `..` signifies going up a folder. In this case, `../` will signify the `Exercises` folder. To save an object in the `IntroToR` folder, say, we therefore write `../../`.

##### Read/Write

Several nifty functions exist to read and write files.

```{r, eval = F}
?read.csv() #reads .csv files into data.frames
?read.table()
?readRDS()
```

```{r, eval = F}
?write.csv()
?write.table()
?saveRDS()
```

Many packages also allow you to read/write different files. Technically `read.csv()` comes from the `utils` package, which is installed directly with R.

```{r, eval = F}
?readr::read_csv()
?terra::rast() #in spatial ecology we use the Terra package to read spatial data
```

##### Data frames

Data frames are the easiest ways to manipulate, analyse, and visualize data in `.R`. For these examples we will use data from the package `palmerpenguins`, a toy data set for exploration and visualization. See <https://allisonhorst.github.io/palmerpenguins/>.

```{r, eval = F}
#install.packages("Rtools")
install.packages("palmerpenguins")
library(palmerpenguins)
```

```{r, eval = F}
# Let us first call and save a copy of the PP dataset penguins
df <- penguins
```

###### Indexing

We have seen ways to index data frames in `./02DataTypes.Rmd`. As a refresher, we will show ways to index the palmer penguins data set.

```{r, eval = F}
df[1] # first row
```

```{r, eval = F}
df[1,1] #first row and first column
```

```{r, eval = F}
df[1,] # first row
```

```{r, eval = F}
df[,1] # first column
```

```{r, eval = F}
df$species #species column
```

```{r, eval = F}
df[, "island"] # named column
```

###### Subset-ing

Subset-ing is very similar to indexing, and usually combines indexing with logical operators.

```{r, eval = F}
df[df$species == "Adelie",] #rows where the species is Adelie
```

```{r, eval = F}
df[df$sex == "female" & df$year > 2006, ]
```

`:` is used as a "through" operator. I.e. `3:6` represents `3, 4, 5, 6`. We can use this to subset our data frame.

```{r, eval = F}
df[1:3, 1:3]
```

###### Manipulating

We can apply `mean()`, `sum()`, etc.. to data frame columns.

```{r, eval = F}
mean(df$bill_len)
```

Note we get `NA` as an answer. Why is that...? There are Nas in the dataset. We can remove them, or we can use a nice trick of `mean():`

```{r, eval = F}
mean(df$bill_len, na.rm = TRUE)
```

To identify `NA`s, we use base R function `is.na()`.

```{r, eval = F}

#removing NAs uses
df_noNa <- df[!is.na(df$bill_len), ]
mean(df_noNa$bill_len)
```

**Bonus!** What is "wrong" with `df_noNa` ?

Next we can go in and change values! Let's say all the `NA`s in `bill_len` actually should be 0s.

```{r, eval = F}
df[is.na(df$bill_len), "bill_len"] <- 0
```

**Bonus!** where did these changes save?

```{r, eval = F}
mean(df$bill_len)
```

###### Adding Data

We can add columns and rows to the data!

```{r, eval = F}
df$new_col <- NA #creates a new column with NA values
```

This can also be done using functions

```{r, eval = F}
?cbind()
cbind(df, df$species) #adding the species column to the df
```

**Bonus!** Did the above snippet of code modify `df`?

###### Apply

`apply()` refers to a suite of base R functions which apply a function over a vector, array, or list. These have include `lapply()` , `tapply()`, `sapply()`, `mapply()`.

```{r, eval = F}
# Here we subset df to numerical columns (which are cols. 3 to 6, where : indicates through). We apply the function mean over the columns, and remove Nas.
apply(df[, 3:6], 2, mean, na.rm = TRUE)
```

###### Example R/W

Finally, lets write our toy data to the data folder as a `.csv` file.

```{r, eval = F}
df <- penguins #lets overwrite df with the penguin data to save the original

write.csv(df, "./exercises/exercises_module1/data/penguins.csv" , row.names = F)
```

To make sure we did it right, we will read the data we saved.

```{r}
read.csv("./exercises/exercises_module1/data/penguins.csv")
```

#### Exercises

These can be done in the `.Rmd` or in a separate `.R` script. To start, install and load the `gapminder` package.

##### 1. Write a data processing snippet to include only the data points collected after 1995 in Asian countries and save a CSV file

```{r, eval = F}

```

##### 2. Separate the `gapminder` data frame into 5 individual data frames, one for each continent. Store those 5 data frames as an `RData` file in the `results` folder called `continents.RData`.

```{r, eval = F}

```

##### 3. Finish exploring the `gapminder` data frame and a) Find the number of rows and the number of columns, b) Print the data type of each column, c) Explain the meaning of everything that `str(gapminder)` prints

```{r, eval = F}

```

##### 4. In which years has the GDP of Canada been larger than the average of all data points?

```{r, eval = F}

```

##### 5. Find the mean life expectancy of Switzerland before and after 2000

```{r, eval = F}

```

##### 6. You discovered that all the entries from 2007 are actually from 2008. Create a copy of the full `gapminder` data frame in an object called `gp`. Then change the year column to correct the entries from 2007.

```{r, eval = F}

```

##### **Bonus -** Find the mean life expectancy and mean gdp per continent using the function `tapply`

hint: to understand the tapply function, use the `?`

```{r, eval = F}

```

#### Solutions

```{r, eval = F}
#install.packages(gapminder)
library(gapminder)

```

##### 1. Write a data processing snippet to include only the data points collected after 1995 in Asian countries and save a CSV file

```{r,eval=F}
asia<-gapminder[gapminder$year > 1995 & gapminder$continent=="Asia", ]

write.table(asia,
            file = "data/gapminder_after1995_asia.csv",
            sep = ",", 
            quote = FALSE, 
            row.names = FALSE)
```

##### 2. Separate the `gapminder` data frame into 5 individual data frames, one for each continent. Store those 5 data frames as an `RData` file in the `results` folder called `continents.RData`.

```{r, eval = F}

asia<-gapminder[gapminder$continent=="Asia", ]
africa<-gapminder[gapminder$continent=="Africa", ]
oceania<-gapminder[gapminder$continent=="Oceania", ]
europe<-gapminder[gapminder$continent=="Europe", ]
americas<-gapminder[gapminder$continent=="Americas", ]

save(asia,africa,oceania,europe,americas,file="../results/continents.RData")
```

##### 3. Finish exploring the `gapminder` data frame and a) Find the number of rows and the number of columns, b) Print the data type of each column, c) Explain the meaning of everything that `str(gapminder)` prints

```{r, eval = F}
dim(gapminder)

typeof(gapminder$country)
typeof(gapminder$continent)
typeof(gapminder$year)
typeof(gapminder$lifeExp)
typeof(gapminder$pop)
typeof(gapminder$gdpPercap)

str(gapminder)
```

##### 4. In which years has the GDP of Canada been larger than the average of all data points?

```{r, eval = F}
canada<-gapminder[gapminder$country=="Canada",]
mgdp<-mean(canada$gdpPercap)
canada[canada$gdpPercap>mgdp,"year"]
```

##### 5. Find the mean life expectancy of Switzerland before and after 2000

```{r, eval = F}
swiss<-gapminder[gapminder$country=="Switzerland",]
mean(swiss[swiss$year<2000,]$lifeExp) # Before
mean(swiss[swiss$year>2000,]$lifeExp) # After
```

##### 6. You discovered that all the entries from 2007 are actually from 2008. Create a copy of the full `gapminder` data frame in an object called `gp`. Then change the year column to correct the entries from 2007.

```{r, eval = F}
gp<-gapminder
gp[gp$year==2007,"year"]<-2008
gp[gp$year==2008,]
```

##### **Bonus -** Find the mean life expectancy and mean gdp per continent using the function `tapply`

hint: to understand the tapply function, use the `?`

```{r, eval = F}
?tapply()
```

```{r, eval = F}
tapply(gapminder$lifeExp,gapminder$continent,mean)
```



### 5. Advanced data manipulation

The fourth module participants will familiarize with the dplyr syntax; create pipes with the operator `%>%`; perform operations on data frames using dplyr and tidyr functions; learn how to join columns and rows of different data frames; and implement functions from external packages by reading their documentation in R.

**Module content:**
-   Handling data frames with dplyr
-   Other useful packages
-   Hands-on: advanced data manipulation

[**Script**](./exercises/exercises_module1/scripts/05AdvancedDataManipulation.Rmd)

#### Examples

Here we go over the material.

```{r, eval = F}
# install.packages("dplyr")
library(dplyr)
library(tidyr)
library(palmerpenguins)

df <- penguins
```

##### Dplyr

```{r, eval = F}
?dplyr
browseVignettes(package = "dplyr") # note this opens a window
```

`dplyr` is a package widely used in R analysis, and lets us easily analyse large datasets. It both contains useful functions, and a "grammar" called piping which can be applied to these functions.

Piping uses the `%>%` operator. In short, this operator "passes" the previous lines of code into the next.

Let us look at the following example:

```{r, eval = F}
df %>% 
  dplyr::select(species) #select is a dplyr function, and lets you select columns using names and types. The pipe passes df as an input to select.

# alternatively, one could do select(df, species)
```

Note the `packagName::functionName`. This is good practice for two reasons: 1) trace-ability, and 2) some packages overwrite functions.

Let us try more functions. Common, and useful, functions include `select()`, `filter()`, `mutate()`, `group_by()`, and `summarise()`. Piping becomes more useful once you start to chain more than one function; the output of the previous function is passed onto the next.

```{r, eval = F}

df %>% 
  # select specified columns
  dplyr::select(species, island, year) %>% 
  
  # keeps data where years greater than 2007
  dplyr::filter(year > 2007) %>% 
  
  # adds a column sorting islands into Torgersen and others
  dplyr::mutate(Torgersen = ifelse(island == "Torgersen", TRUE, FALSE)) %>% 
  
  # create groups which are a combination of the values in species and Torgersen
  dplyr::group_by(species, Torgersen) %>%
  
  # Count the number of entries in each group
  dplyr::summarise(number = dplyr::n())
```

###### Mutating joins

Mutating joins are a group of `dplyr` functions which allow you to join two datasets together. Different joins exist: `inner_join()`, `left_join()`, `right_join()`, and `full_join()`.

In the following example, we will create a new data frame with a code for each species, and join it to `df`.

```{r, eval = F}
# first we look at what species are present in the dataset
df %>% select(species) %>% unique(.)
```

```{r, eval = F}
# then we create the code data frame
df_code <- df %>% 
  dplyr::select(species) %>%
  dplyr::mutate(code = case_when(
    species == "Adelie" ~ 1,
    species == "Gentoo" ~ 2,
    species == "Chinstrap" ~ 3
  ))

head(df_code)
```

```{r, eval = F}
# Now we join the two together.
df %>%
  dplyr::left_join(., df_code, by = "species")
```

##### Other Packages

###### Stringr

For fun, lets look at and inplement `stringr`, a string manipulation package.

```{r, eval = F}
# install.packages("stringr")
library(stringr)
```

We will extract the first three letters of each island to create an island code. To do this, we will use

```{r, eval = F}
df %>%
  dplyr::mutate(island_code = stringr::str_sub(island, 1, 3))
```

###### Tidyr

`tidyr` is a useful package that helps clean/sort/manipulate data.

Useful functions include `pivot_longer()` and `pivot_wider()` which are often used in combination with `ggplot2` to plot data (don't worry we will see plotting in the next module!)/

Let's use `pivot_wider()` as an example. These functions can also be piped.

```{r, eval = F}
df %>% 
  pivot_wider(names_from = island, values_from = species)
```

Note! doing the above really isn't useful in this situation but there are times where it is useful to have a wider data-frame. For example, in ecology, this data set can be adapted using pivot_wider to create a site by species data frame.

#### Exercises

You can do these in the notebook or in a separate `.R` script. We again use the `gapminder` library. If it is not installed, install it, and if it is not loaded, load it. Note! If you use a separate `.R` script, you must load the necessary additional libraries.

```{r, eval = F}

```

##### 1. Write one command (can span multiple lines) using pipes that will output a data frame that has only the columns `lifeExp`, `country` and `year` for the records before the year 2000 from African countries, but not for other Continents.

```{r, eval = F}

```

##### 2. Calculate the average life expectancy per country. Which country has the longest average life expectancy and which one the shortest average life expectancy?

```{r, eval = F}

```

##### 3. In the previous hands-on you discovered that all the entries from 2007 are actually from 2008. Write a command to edit the data accordingly using pipes. In the same command filter only the entries from 2008 to verify the change.

```{r, eval = F}

```

##### 4. Lets add data from a different .csv to our data frame.

###### a) Read in the `co2_pcap_cons.csv` found [here](./exercises/exercises_module1/data/co2_pcap_cons.csv).

```{r, eval = F}

```

###### b) Look at the data. What do you notice? What will you need to change to be able to join the two dataframes together?

###### c) Perform the changes you identified in (b)

hint: use `stringr`!

```{r, eval = F}

```

###### d) join both data frames

```{r, eval = F}

```

#### Solutions

```{r, eval = F}
library(gapminder)
library(dplyr)
library(tidyr)
```

##### 1. Write one command ( can span multiple lines) using pipes that will output a data frame that has only the columns `lifeExp`, `country` and `year` for the records before the year 2000 from African countries, but not for other Continents.

```{r, eval = F}
tidy_africa <- gapminder %>%
                dplyr::filter(continent == "Africa") %>%
                dplyr::select(year, country, lifeExp)
head(tidy_africa)
```

##### 2. Calculate the average life expectancy per country. Which country has the longest average life expectancy and which one the shortest average life expectancy?

```{r, eval = F}
gapminder %>%
   dplyr::group_by(country) %>%
   dplyr::summarize(mean_lifeExp = mean(lifeExp)) %>%
   dplyr::filter(mean_lifeExp == min(mean_lifeExp) | mean_lifeExp == max(mean_lifeExp))
```

##### 3. In the previous hands-on you discovered that all the entries from 2007 are actually from 2008. Write a command to edit the data accordingly using pipes. In the same command filter only the entries from 2008 to verify the change.

```{r, eval = F}
gapminder %>%
  dplyr::mutate(year = ifelse(year==2007,2008,year)) %>%
  dplyr::filter(year==2008) %>%
  head()
```

##### 4. Lets add data from a different .csv to our data frame.

###### a) Read in the `co2_pcap_cons.csv` found [here](./exercises/exercises_module1/data/co2_pcap_cons.csv).

```{r, eval = F}
new_data <- read.csv("./exercises/exercises_module1/data/co2_pcap_cons.csv")
```

###### b) Look at the data. What do you notice? What will you need to change to be able to join the two dataframes together?

-   Change the data frame such that `country` , `year`, and `emissions` are separate columns.

-   Change `year` to numeric.

###### c) Perform the changes you identified in (b)

hint: use `stringr`!

```{r, eval = F}
new_data_t <- new_data %>% 
  pivot_longer(cols = -country, names_to = "year", values_to = "emissions") %>%
  mutate(year = as.numeric(str_extract(year, "\\d+")))

```

Note here i use `str_extract()` combined with regex, but you can use `str_sub()` with `start = 2` and `end = 5` as seen in the examples above!

###### d) join both data frames

```{r, eval = F}
gapminder %>% left_join(., new_data_t, by = c("country", "year"))
```



### 6. Generating visual outputs

This section will show participants how to (1) Create basic plots using base R functions; (2) Understand how to connect data frames with ggplot2; (3) create basic graphs with ggplot2; (4) use factors to customize graphics in ggplot2; (5) use RMarkdown to generate customized reports.

**Module content:**
-   Figures with base R
-   Graphics with `ggplot2`

[**Script**](./exercises/exercises_module1/scripts/06GeneratingOutputs.Rmd)

#### Examples

Here we go over the material.

We will continue using the palmer penguins as an example.

```{r, eval = F}
library(ggplot2)
library(palmerpenguins)
library(tidyr)
library(dplyr)

df <- penguins
```

##### Basic Plots

R has a number of basic plots. The most basic of these functions is `plot()`.

```{r, eval = F}
plot(df$species, df$bill_length_mm)
```

```{r, eval = F}
plot(df$bill_depth_mm, df$bill_length_mm)
```

A similar function exists to look at distributions: `hist()`

```{r, eval = F}
hist(df$body_mass_g)
```

##### Using `ggplot2`

Most visualizations, however, are done with the package `ggplot2`. This package allows for more control over the graphs, and produces better visualizations. Furthermore, `ggplot2` is tailored to making visuals from data frames.

In this section, we go over the basics of `ggplot2`, but <https://ggplot2.tidyverse.org/> has a handy cheatsheet!

The basic architecture of a ggplot2 plot is `ggplot(df, aes(x, y)) + ...`, where `df` represents the data frame of data, `x` your x variable, and `y` your y variable. `aes` stands for aesthetic. The type of plot is specified after the first call, for example a scatter plot is created via `ggplot(df, aes(x, y)) + geom_point()`.

Let's first set up a color scheme:

```{r, eval = F}
colorcode <- c("darkorange","purple","cyan4") #scheme from citation(palmerpenguins)
```

Now we look at the number of penguins of each species on each island. This calls for a bar chart; in ggplot you do not specify a `y` variable for a bar chart as this is automatically calculated. Instead, you supply a `fill` aesthetic.

```{r, eval = F}

df %>%
  ggplot(aes(x = island, fill = species))+
  geom_bar()+
  scale_fill_manual(values = colorcode, 
                    guide = NULL) +
  # basic theme
  theme_minimal()

```

If we want to separate by species, we can use `facet_wrap()`:

```{r, eval = F}
df %>% 
  ggplot(aes(x = island, fill = species))+ 
  geom_bar()+ 
  scale_fill_manual(values = colorcode, guide = NULL) + 
  theme_minimal()+
  facet_wrap(.~species)
```

Note that the same plot can be created a different way, specifying the `y` aesthetic. Notice the differences!

```{r, eval = F}
df %>% 
  
  # we calculate the number of penguins of each species at each island
  group_by(island, species) %>%
  summarise(n = n()) %>%
  
  #now we can specify n
  ggplot(aes(x = island, y = n, fill = species))+ 
  
  # since we specify n, we need to tell the function via stat = "identity"
  geom_bar(stat = "identity")+ 
  scale_fill_manual(values = colorcode, guide = NULL) + 
  theme_minimal()+
  facet_wrap(.~species)
```

Lets look at bill length in each species.

```{r, eval = F}

df %>%
  ggplot(aes(x = species, y = bill_length_mm, fill = species ))+
  geom_boxplot()+
  scale_fill_manual(values = colorcode, guide = NULL) + 
  theme_minimal()

```

Now lets plot the bill length and depth against each other:

```{r, eval = F}
df %>% 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species, shape = island))+ 
  geom_point()+ 
  scale_color_manual(values = colorcode, guide = NULL) + 
  theme_minimal()
```

Now Lets look at how many penguins of each sex and species there are using a bar chart.

```{r, eval = F}

df %>% 
  ggplot(aes(x = sex, fill = species)) +
  
  # alpha corresponds to transparency. the smaller, the more transparent
  geom_bar(alpha = 0.8) +
  
  # code to control the 'fill' aes
  scale_fill_manual(values = colorcode, 
                    guide = NULL) +
  
  # basic theme
  theme_minimal() +
  
  # bringing multiple 
  facet_wrap(~species, ncol = 1) +
  coord_flip()

# note this snippet of code is adapted from citation("palmerpenguins")
```

Let's look at the distribution of body mass by sex.

```{r, eval = F}

df %>%
  ggplot(aes(x = body_mass_g, fill = species))+
  geom_histogram(alpha = 0.8) +
  scale_fill_manual(values = colorcode, 
                    guide = NULL) +
  theme_minimal() +
  facet_wrap(. ~ sex)

```

If we want to separate all of them out in a nice grid format, we can use use `facet_grid()` :

```{r, eval = F}
df %>%
  ggplot(aes(x = body_mass_g, fill = species))+
  geom_histogram(alpha = 0.8) +
  scale_fill_manual(values = colorcode, 
                    guide = NULL) +
  theme_minimal() +
  facet_grid(species ~ sex)
```

Finally, lets look at this data over time. Specifically, the number of penguins found each year!

```{r, eval = F}

df %>%
  group_by(year, species) %>%
  summarise(n = n()) %>%
  ggplot(aes(x = year, y = n, color = species))+
  geom_point()+
  geom_line()+
  scale_color_manual(values = colorcode, 
                    guide = NULL) +
  theme_minimal()+
  
  # here we specify the x and y axes. We change the x axis limits so that we don't end up with 2007.5 and 2008.5 in the graph.
  xlab("Year")+
  ylab("Number of penguins")+
  scale_x_continuous(breaks = c(2007, 2008, 2009))


```

Many more graphs exist, and many aspects of a graph can be modified in `ggplot2`. The best way to learn is by doing them!

#### Exercises

You can do these in the notebook or in a separate `.R` script.

For these exercises, you need libraries `ggplot2`, `medicaldata`, `dplyr` and `tidyr`.

```{r, eval = F}

```

##### 1. Lets start by looking at the clinics.

###### a) How many clinics participated in the study

```{r, eval = F}

```

###### b) How many valid tests were performed on each one?

```{r, eval = F}

```

###### c) Did the testing trend vary over time?

```{r, eval = F}

```

##### 2. Now Looking at the patients.

###### a) How many patients tested positive vs negative in the first 100 days of the pandemic?

```{r, eval = F}

```

###### b) Do you notice any difference with the age of the patients?

**Hint: You can make two age groups and calculate the percentage each age group in positive vs negative tests.**

```{r, eval = F}

```

##### 3. Processing times

###### a) Look at the specimen processing time to receipt, did the sample processing times improve over the first 100 days of the pandemic?

```{r, eval = F}

  
```

###### b) Plot the median processing times of each day over the course of the pandemic and then compare the summary statistics of the first 50 vs the last 50 days.

```{r, eval = F}

```

##### 4. Bonus - Higher viral loads are detected in less PCR cycles. What can you observe about the viral load of positive vs negative samples. Do you notice anything differences in viral load across ages in the positive samples?

Hint: Also split the data into two age groups and try using geom_boxplot()

```{r, eval = F}

```

#### Solutions

```{r, eval = F}
suppressPackageStartupMessages(library(ggplot2)) 
suppressPackageStartupMessages(library(medicaldata))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
```

Use the `covid_testing` data set and everything you've learned so far to answer the following questions:

##### 1.Lets start by looking at the clinics.

```{r, eval = F}
covid <- medicaldata::covid_testing
```

###### a) How many clinics participated in the study

```{r, eval = F}
clinics<- covid %>%
          dplyr::select(subject_id,clinic_name,result,pan_day) %>%
          dplyr::distinct()
length(unique(clinics$clinic_name))
```

###### b) How many valid tests were performed on each one?

```{r, eval = F}
clinics %>%
  dplyr::filter(result!="invalid") %>%
  dplyr::group_by(clinic_name) %>%
  dplyr::summarize(n_test = length(clinic_name)) %>%
  dplyr::arrange(desc(n_test))
```

###### c) Did the testing trend vary over time?

```{r, eval = F}
covid %>%
  filter(pan_day<=100) %>%
  group_by(pan_day) %>%
  summarize(n=length(result)) %>%
  ggplot(.,aes(x=pan_day,y=n))+
    geom_point()+
    geom_line()+
    ylab("Number of tests per day")+
    xlab("Pandemic day")
```

##### 2. Now Looking at the patients.

###### a) How many patients tested positive vs negative in the first 100 days of the pandemic?

```{r, eval = F}
covid %>%
  filter(result!="invalid" & pan_day<=100) %>%
  group_by(result) %>%
  summarize(n=length(subject_id))
```

###### b) Do you notice any difference with the age of the patients?

**Hint: You can make two age groups and calculate the percentage each age group in positive vs negative tests.**

```{r, eval = F}
tsts_age<-covid %>%
            filter(result!="invalid" & pan_day<=100) %>%
            mutate(age_group=ifelse(age<=21,"children","adults")) %>%
            group_by(age_group,result) %>%
            summarize(n=length(subject_id)) %>%
            mutate(percent_total=n/sum(n)*100)
tsts_age
```

##### 3. Processing times

###### a) Look at the specimen processing time to receipt, did the sample processing times improve over the first 100 days of the pandemic?

```{r, eval = F}
covid %>%
  group_by(pan_day) %>%
  dplyr::summarise(median_col_rec_tat=median(col_rec_tat)) %>%
  ggplot(.,aes(x=pan_day,y=median_col_rec_tat)) +
  geom_point()+
  geom_line()
```

###### b) Plot the median processing times of each day over the course of the pandemic and then compare the summary statistics of the first 50 vs the last 50 days.

```{r, eval = F}
covid %>%
  mutate(pan_day_group=ifelse(pan_day<50,"first_50","last_50")) %>%
  group_by(pan_day_group) %>%
  dplyr::summarise(mean_col_rec_tat=mean(col_rec_tat),
                   median_col_rec_tat=median(col_rec_tat),
                   min_col_rec_tat=min(col_rec_tat),
                   max_col_rec_tat=max(col_rec_tat))
```

##### 4. Bonus - Higher viral loads are detected in less PCR cycles. What can you observe about the viral load of positive vs negative samples. Do you notice anything differences in viral load across ages in the positive samples?

Hint: Also split the data into two age groups and try using geom_boxplot()

```{r,eval = F, warning=FALSE}
ggplot(covid,aes(y=ct_result,x=result,color=result))+
  geom_boxplot()
```

```{r,eval = F, warning=FALSE}
covid %>%
  mutate(age_group=ifelse(age<=21,"children","adults")) %>%
  ggplot(.,aes(y=ct_result,x=result,color=age_group))+
    geom_boxplot()
```



### Repo structure
```
├── README.md
├── .gitignore
├── _config.yaml
├── Intro_Mon-2510.Rproj
├── Exercises
│   └── README.md
│   ├── data
│   │   └── co2_pcap_cons.csv
│   │   └── penguins.csv
│   ├── scripts
│   │   └── 01GettingStarted.Rmd and .html
│   │   └── 02DataTypes.Rmd and .html
│   │   └── 03ControlStructuresAndFunctions.Rmd and .html
│   │   └── 04BasicDataManipulation.Rmd and .html
│   │   └── 05AdvancedDataManipulation.Rmd and .html
│   │   └── 06GeneratingOutputs.Rmd and .html
├── Slides
│   └── IntroToR_CBHWorkshop.pptx
└── Outline
    └── workshopoutline.pdf #using the template provided
```

### References
The materials for this workshop were based on the following sources:
-   [Base R Cheat Sheet](https://iqss.github.io/dss-workshops/R/Rintro/base-r-cheat-sheet.pdf)\
-   [Google's R Style Guide](https://google.github.io/styleguide/Rguide.html)
-   [Mastering Software Development in R](https://bookdown.org/rdpeng/RProgDA/)

Data in this workshop comes from: 
- Horst AM, Hill AP, Gorman KB (2020). palmerpenguins: Palmer Archipelago (Antarctica) penguin data. R package version 0.1.0. https://allisonhorst.github.io/palmerpenguins/. doi: 10.5281/zenodo.3960218
- Gapminder data free from www.gapminder.org, CC-BY LICENSE
- Bryan J (2025). _gapminder: Data from Gapminder_. doi:10.32614/CRAN.package.gapminder <https://doi.org/10.32614/CRAN.package.gapminder>, R package version 1.0.1, <https://CRAN.R-project.org/package=gapminder>.
- Higgins P (2021). _medicaldata: Data Package for Medical Datasets_. doi:10.32614/CRAN.package.medicaldata <https://doi.org/10.32614/CRAN.package.medicaldata>, R package version 0.2.0, <https://CRAN.R-project.org/package=medicaldata>.

*Workshop created as part of the McGill Initiative in Computational Medicine*

