# (PART) Modules {-}

# Practical 1

## Requirements
- R and RStudio must be installed prior to the session - you can follow the instructions below
  - [Install R](https://cran.r-project.org/)
  - [Install RStudio](https://posit.co/download/rstudio-desktop/)
  
## Lecture
<iframe 
  src="https://docs.google.com/presentation/d/10LFAUpZFSsJ2ZsJH6y70J7RfVGb97OfR/preview" 
  width="640" 
  height="480" 
  frameborder="0" 
  allow="autoplay">
</iframe>

<!--<iframe width="640" height="360" src="YOUTUBE EMBED LINK" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen></iframe>
-->

## Lab

Materials for this practical can be downloaded [from here](https://github.com/MJousse/IntroToR/).

### 1. R basics
In the first module of the workshop, the goals are to familiarize with the language and the logic behind it; Get started with R studio and create your first project; Create your first `.R` file to write down the live code; and get comfortabble with installing packages.

**Module content:**
-   Seeking help
-   Installing packages

[**Script**](./exercises/exercises_module1/scripts/01GettingStarted.Rmd)

#### R Markdown File

R Markdown files act as notebooks which can make it easier to document and share code. All the results from the code as well as your comments and descriptions can be combined into one, tidy document. This makes it easy to combine insights, protocols and figures in a practical and organized fashion.

#### Setting up:

##### Installing a package

To install a package, we simply use the function `install.packages()` with the name of the desired package in quotation marks (i.e. as a string!)

```{r, eval=F}
install.packages("packageName1")

# for multiple packages:
install.packages(c("packageName1", "packageName2"))

```

##### Loading packages

To load a package you simply use the `library()` function. Note function takes the name of an installed library without quotation marks.

```{r, eval=F}

library(packageName1)

```

#### R Basics:

##### Variable names

There are several naming conventions in coding: `snake_case` and `CamelCase` are the most common. There is no particular preference in R for naming, but make sure you are **consistent** and that your variable names are **meaningful**.

##### Case sensitivity

R is case sensitive. This means that `var1` and `Var1` represent two different objects.

##### 1 based indexing

Indexing (calling elements in a list) in R starts with 1. The first element has index 1, the second 2, etc..

```{r}
# Initialising a vector
var1 <- c('a', 'b', "c", "d") # note ' and " are the same in R :)

# Here we call elements 1 and 2 of var1
var1[1]
var1[2]

#Now we look at how long var1 is using the length() function

length(var1)

```

### Exercises

#### 1. Create a file

Create a new .R script in R studio. Save it as `Script01` in the `IntroToR` folder.

#### 2. Install base packages

In `Script01`, install the following packages: `"data.table"`,`"datasets"`,`"devtools"`,`"dplyr"`,`"ggplot2"`,`"plyr"`,`"medicaldata"`,`"gapminder"`,`"RColorBrewer"`,`"rmarkdown"`,`"stringr"`,`"tidyr"`,`"tidyverse"`,`"viridis"`.

#### 3. Load a package

Load all packages.

### Solutions

#### 1. Creating a file

#### 2. Install base packages

```{r, eval = F}
install.packages(c("data.table","datasets","devtools","dplyr","ggplot2","plyr","medicaldata","gapminder","RColorBrewer","rmarkdown","stringr","tidyr","tidyverse","viridis"))
```

#### 3. load packages

```{r, eval = F}
library(data.table)
library(datasets)
library(devtools)
library(dplyr)
library(ggplot2)
library(plyr)
library(medicaldata)
library(gapminder)
library(RColorBrewer)
library(rmarkdown)
library(stringr)
library(tidyr)
library(tidyverse)
library(viridis)
```





### 2. Data types: attributes and built-in functions

In this section participants will understand the differences between classes, objects and data types in R; Compute arithmetic operations; Use logical operators; create objects of different types, learn about their attributes and apply some built-in functions in R; Subset and index objects; and get comfortable with vectorized operations.

**Module content:**
-   Vectors
-   Lists
-   Factors
-   Data frames
-   Arrays
-   Coercion

[**Script**](./exercises/exercises_module1/scripts/02DataTypes.Rmd)

### 3. Control structures
In this module, participants learn to use if-else, while, and for loops; as well as building their own functions; and what packages are.

**Module Content**
* If..Else
* While loop
* For loop
* Functions
* Packages

[**Script**](./exercises/exercises_module1/scripts/03ControlStructuresAndFunctions.Rmd)

### 4. Basic data manipulation

In this module participants will learn how to read/write data to/from files with different formats (.tsv, .csv); become familiar with basic data-frame operations; (3) index and subset data frames using base R; manipulate individual data frame columns.

**Module content:**
-   Reading/writing data
-   Exploring data frames

[**Script**](./exercises/exercises_module1/scripts/04BasicDataManipulation.Rmd)

### 5. Advanced data manipulation

The fourth module participants will familiarize with the dplyr syntax; create pipes with the operator `%>%`; perform operations on data frames using dplyr and tidyr functions; learn how to join columns and rows of different data frames; and implement functions from external packages by reading their documentation in R.

**Module content:**
-   Handling data frames with dplyr
-   Other useful packages
-   Hands-on: advanced data manipulation

[**Script**](./exercises/exercises_module1/scripts/05AdvancedDataManipulation.Rmd)

### 6. Generating visual outputs

This section will show participants how to (1) Create basic plots using base R functions; (2) Understand how to connect data frames with ggplot2; (3) create basic graphs with ggplot2; (4) use factors to customize graphics in ggplot2; (5) use RMarkdown to generate customized reports.

**Module content:**
-   Figures with base R
-   Graphics with `ggplot2`

[**Script**](./exercises/exercises_module1/scripts/06GeneratingOutputs.Rmd)

### Repo structure
```
├── README.md
├── .gitignore
├── _config.yaml
├── IntroToR.Rproj
├── Exercises
│   └── README.md
│   ├── data
│   │   └── co2_pcap_cons.csv
│   │   └── penguins.csv
│   ├── scripts
│   │   └── 01GettingStarted.Rmd and .html
│   │   └── 02DataTypes.Rmd and .html
│   │   └── 03ControlStructuresAndFunctions.Rmd and .html
│   │   └── 04BasicDataManipulation.Rmd and .html
│   │   └── 05AdvancedDataManipulation.Rmd and .html
│   │   └── 06GeneratingOutputs.Rmd and .html
├── Slides
│   └── IntroToR_CBHWorkshop.pptx
└── Outline
    └── workshopoutline.pdf #using the template provided
```

### References
The materials for this workshop were based on the following sources:
-   [Base R Cheat Sheet](https://iqss.github.io/dss-workshops/R/Rintro/base-r-cheat-sheet.pdf)\
-   [Google's R Style Guide](https://google.github.io/styleguide/Rguide.html)
-   [Mastering Software Development in R](https://bookdown.org/rdpeng/RProgDA/)

Data in this workshop comes from: 
- Horst AM, Hill AP, Gorman KB (2020). palmerpenguins: Palmer Archipelago (Antarctica) penguin data. R package version 0.1.0. https://allisonhorst.github.io/palmerpenguins/. doi: 10.5281/zenodo.3960218
- Gapminder data free from www.gapminder.org, CC-BY LICENSE
- Bryan J (2025). _gapminder: Data from Gapminder_. doi:10.32614/CRAN.package.gapminder <https://doi.org/10.32614/CRAN.package.gapminder>, R package version 1.0.1, <https://CRAN.R-project.org/package=gapminder>.
- Higgins P (2021). _medicaldata: Data Package for Medical Datasets_. doi:10.32614/CRAN.package.medicaldata <https://doi.org/10.32614/CRAN.package.medicaldata>, R package version 0.2.0, <https://CRAN.R-project.org/package=medicaldata>.

*Workshop created as part of the McGill Initiative in Computational Medicine*

