# Practical 3

## Requirements

The `tidyverse` and `nnet` packages are required for this section and can be installed with:

`install.packages(c("tidyverse","nnet"))`

You will also need previous experience with probability to best benefit from the materials.

## Lecture

<iframe 
  src="https://docs.google.com/presentation/d/1_n8ycxON2l7g1Wo_OOEVX1oXYNI-KNxt/preview" 
  width="640" 
  height="480" 
  frameborder="0" 
  allow="autoplay">
</iframe>

<!--<iframe width="640" height="360" src="YOUTUBE EMBED LINK" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen></iframe>
-->

## Lab

### Linear regression

For this part of the practical, you can download this [**script**](./exercises/exercises_module3/scripts/linear_regression.Rmd).


#### Loading libraries
```{r Loading libraries linear regr, eval = F}
library(tidyverse)
```

#### Verifying the dataset
Let's take the iris data set, and check if the petal length and width follow a linear correlation
```{r plot data linear regr, eval = F}
ggplot(iris) + geom_point(aes(Petal.Length, Petal.Width, col = Species))
cor(iris$Petal.Length, iris$Petal.Width)
```
The data seems to be linearly correlated. The species have different ranges of sizes, but they seem to all follow the same trend.


#### Fitting with MLE
##### 1- Estimating beta1
$$\hat{\beta_1} = \frac{\sum^{N}_{i=1}(x_{i} - \bar{x})(y_{i} - \bar{y})}{\sum_{i=1}^{N}(x_{i} - \bar{x})^{2}} = \frac{SS_{xy}}{SS_{xx}}$$

```{r beta1 estimation linear regr, eval = F}
mean_x = mean(iris$Petal.Length)
mean_y = mean(iris$Petal.Width)
numerator = sum((iris$Petal.Length - mean_x) * (iris$Petal.Width - mean_y))
denominator = sum((iris$Petal.Length - mean_x)**2)
beta1 = numerator / denominator
```
##### 2- Estimating beta0
$$\hat{\beta_0} = \bar{y} - \hat{\beta_{1}}\bar{x}$$
```{r beta0 estimation linear regr, eval = F}
beta0 = mean_y - beta1*mean_x
```
##### 3- Visualize the regression
```{r visualization linear regr, eval = F}
regression_df = data.frame(x = 1:7, y = beta0+c(1:7)*beta1)
ggplot(iris) + geom_point(aes(Petal.Length, Petal.Width, col = Species)) +
            geom_line(aes(x, y), data = regression_df)
```


#### Build-in linear regression
lm is the build-in regression function
```{r lm linear regr, eval = F}
linear_model <- lm(Petal.Width ~ Petal.Length, data = iris)
summary(linear_model)
```
Plot the results to compare with the MLE
```{r plot lm linear regr, eval = F}
regression_df2 = data.frame(x = 1:7, y = linear_model$coefficients[1]+c(1:7)*linear_model$coefficients[2])
ggplot(iris) + geom_point(aes(Petal.Length, Petal.Width, col = Species)) +
  geom_line(aes(x, y), data = regression_df2)
```


#### Predicting results
Say we remove the versicolor species during model construction,  can we predict the Petal Width in function of its Length?

1- We remove the versicolor data and compute the model
```{r pred 1 linear regr, eval = F}
iris_subset <- iris %>% filter(Species != 'versicolor')
iris_model <- lm(Petal.Width ~ Petal.Length, data = iris_subset)
```
2- We extract the versicolor Petal Length
```{r pred 2.1 linear regr, eval = F}
versicolor_p <- iris %>% filter(Species == 'versicolor') %>%
                          mutate(Predicted_pw = iris_model$coefficients[1] + Petal.Length*iris_model$coefficients[2])
```
Alternatively
```{r pred 2.2 linear regr, eval = F}
predict.lm(iris_model, versicolor_p, type = 'response')
```
3- Plot the results
```{r pred 3 linear regr, eval = F}
ggplot(iris) + geom_point(aes(Petal.Length, Petal.Width, col = Species)) +
  geom_point(aes(Petal.Length, Predicted_pw), data = versicolor_p, color = 'darkgreen', shape = '+', size = 5)
```
4- Compute the MSE
```{r pred 4 linear regr, eval = F}
MSE = mean((versicolor_p$Petal.Width - versicolor_p$Predicted_pw)**2)
print(MSE)
```


#### Using multiple variables
What if we also use the Sepal values to predict the Petal Width?

1- We compute the model after removal of versicolor
```{r multi vars 1 linear regr, eval = F}
iris_model2 <- lm(Petal.Width ~ Petal.Length + Sepal.Length + Sepal.Width, data = iris_subset)
```
2- We extract the versicolor Petal Length
```{r multi vars 2 linear regr, eval = F}
versicolor_p2 <- iris %>% filter(Species == 'versicolor') %>%
  mutate(Predicted_pw = iris_model2$coefficients[1] +
                        Petal.Length*iris_model2$coefficients[2] +
                        Sepal.Length*iris_model2$coefficients[3] +
                        Sepal.Width*iris_model2$coefficients[4])
```
3- Plot the results
```{r multi vars 3 linear regr, eval = F}
ggplot(iris) + geom_point(aes(Petal.Length, Petal.Width, col = Species)) +
  geom_point(aes(Petal.Length, Predicted_pw), data = versicolor_p2, color = 'darkgreen', shape = '+', size = 5)
```

4- Compute the MSE
```{r multi vars 4 linear regr, eval = F}
MSE2 = mean((versicolor_p2$Petal.Width - versicolor_p2$Predicted_pw)**2)
print(MSE2)
```
In this case, the MSE is a lot larger than the previous model, so adding more variables did not help. WWhat are the correlations between the variables?
```{r multi vars 5 linear regr, eval = F}
ggplot(iris) + geom_point(aes(Sepal.Width, Petal.Width, col = Species))
cor(iris$Sepal.Width, iris$Petal.Width)
ggplot(iris) + geom_point(aes(Sepal.Length, Petal.Width, col = Species))
cor(iris$Sepal.Length, iris$Petal.Width)
```
Quick test with only the Petal Length and Sepal Length
```{r multi vars 6 linear regr, eval = F}
iris_model3 <- lm(Petal.Width ~ Petal.Length + Sepal.Length, data = iris_subset)
versicolor_p3 <- iris %>% filter(Species == 'versicolor') %>%
  mutate(Predicted_pw = iris_model3$coefficients[1] +
                        Petal.Length*iris_model3$coefficients[2] +
                        Sepal.Length*iris_model3$coefficients[3])
MSE3 = mean((versicolor_p3$Petal.Width - versicolor_p3$Predicted_pw)**2)
ggplot(iris) + geom_point(aes(Petal.Length, Petal.Width, col = Species)) +
  geom_point(aes(Petal.Length, Predicted_pw), data = versicolor_p3, color = 'darkgreen', shape = '+', size = 5)
print(MSE3)
```
This is the worst MSE of the 3, possibly because of overfitting.


#### Hands-on
Create your own linear model

Using mtcars, find the best linear model to predict the miles per gallon (mpg)
Use the first 2/3 of the data as training data for the model and the last third as test data
```{r hands on linear regr, eval = F}
mtcars_train <- mtcars[1:22,]
mtcars_test <- mtcars[23:32,]
```


#### Hands-on solutions

The solutions for this part are in this [**script**](./exercises/exercises_module3/results/linear_regression_answers.Rmd).

Solution
```{r hands on solution, eval = F}
# To help us select a set of variables to use, we can test the correlation of all variables with mpg
mtcars_train %>% apply(2, cor, y=mtcars_train$mpg)
# We have a good negative correlation with the number of cylinders, the displacement,
# horsepower and weight. We will construct a model with those

# 1- Compute the model
cars_model <- lm(mpg ~ cyl+disp+hp+wt, data = mtcars_train)

# 2- Compute the results
cars_p <- mtcars_test %>% 
  mutate(Predicted_mpg = cars_model$coefficients[1] +
                          cyl*cars_model$coefficients[2] +
                          disp*cars_model$coefficients[3] +
                          hp*cars_model$coefficients[4] +
                          wt*cars_model$coefficients[5])

# 3- Compute the MSE
MSE = mean((cars_p$mpg - cars_p$Predicted_mpg)**2)
```
Note: this may not be the overall best solution, it's only one example



### Logistic regression

For this part of the practical, you can download this [**script**](./exercises/exercises_module3/scripts/logistic_regression.Rmd).


#### Loading libraries
```{r Loading libraries logistic regr, eval = F}
library(tidyverse)
library(nnet)
```


#### Predicting species with a linear model
We remove the versicolor species to only have two species
```{r lm 1 logistic regr, eval = F}
iris_subset <- iris %>% filter(Species != 'versicolor')
```
Predict the species based on the petal dimensions
```{r lm 2 logistic regr, eval = F}
iris_subset$train_test <- sample(c('train', 'test'), nrow(iris_subset), replace = TRUE, prob = c(0.8, 0.2))
iris_model <- lm(Species ~ Petal.Length + Petal.Width, data = iris_subset[iris_subset$train_test == 'train',])
predictions <- predict.lm(iris_model, iris_subset[iris_subset$train_test == 'test',], type = 'response')
```
Convert the predictions to species and compare to the truth
```{r lm 3 logistic regr, eval = F}
table(round(predictions), iris_subset[iris_subset$train_test == 'test', 'Species'])
```


#### Predicting 2 species with a log model
```{r log 2 categories logistic regr, eval = F}
iris_log_model <- glm(Species ~ Sepal.Length +  Sepal.Width,
                  family = 'binomial',
                  data = iris_subset[iris_subset$train_test == 'train',])

predictions_log <- predict.glm(iris_log_model,
                               iris_subset[iris_subset$train_test == 'test',],
                              type = 'response')

predictions_log <- ifelse(predictions_log >= 0.5,"virginica", "setosa")

table(predictions_log, iris_subset[iris_subset$train_test == 'test', 'Species'])
```


#### Predicting 3 species with a log model
```{r log 3 categories 1 logistic regr, eval = F}
iris$train_test <- sample(c('train', 'test'), nrow(iris), replace = TRUE, prob = c(0.8, 0.2))
iris_train <- iris %>% filter(train_test == 'train')
iris_test <- iris %>% filter(train_test == 'test')

iris_multinomial <- multinom(Species ~ Sepal.Length +  Sepal.Width,
                      data = iris_train)

predictions_multi <- predict(iris_multinomial,
                               iris_test,
                               'prob')
```
Convert the probabilities to species with a softmax
```{r log 3 categories 2 logistic regr, eval = F}
species_predictions <- predictions_multi %>%
  apply(1, function(x) return(c('setosa', 'versicolor', 'virginica')[which.max(x)]))

table(species_predictions, iris_test$Species)
```


#### Hands-on
From the diamonds model, predict the cut

First divide the data in training and testing datasets
```{r hands on logistic regr, eval = F}
diamonds$train_test <- sample(c('train', 'test'), nrow(diamonds), replace = TRUE, prob = c(0.8, 0.2))
diamonds_train <- diamonds %>% filter(train_test == 'train')
diamonds_test <- diamonds %>% filter(train_test == 'test')
```


#### Hands on solution

The solutions for this part are in this [**script**](./exercises/exercises_module3/results/logistic_regression_answers.Rmd).

Solution
```{r hands on solution logistic regr, eval = F}
diamonds_multinomial <- multinom(cut ~ carat + color + clarity + depth,
                             data = diamonds_train)

predictions_diamonds <- predict(diamonds_multinomial,
                             diamonds_test,
                             'prob')

# Convert the probabilities to categories
categories_predictions <- predictions_diamonds %>%
  apply(1, function(x) return(colnames(predictions_diamonds)[which.max(x)]))

table(categories_predictions, diamonds_test$cut)
```
Note: this may not be the overall best solution, it's only one example



### Statistical testing

For this part of the practical, you can download this [**script**](./exercises/exercises_module3/scripts/testing.Rmd).


#### Loading libraries
```{r Loading libraries stats test, eval = F}
library(tidyverse)
```

#### Two-samples t-test
1- First, we create two random samples from Normal distributions
```{r ttest 1 stats test, eval = F}
set.seed(20221011) # Seed for reproducibility
n <- 100 # Sample size

mu1 <- 0 # Mean of sample 1
mu2 <- 0.1 # Mean of sample 2

sd1 <- 1 # Standard deviation 1
sd2 <- 1 # Standard deviation 2
```
Generate two samples of of size n of normally-distributed data.

Sample x1 has a mean of mu1 and standard deviation sd1

Sample x2 has a mean of mu2 and standard deviation sd2
```{r ttest 2 stats test, eval = F}
x1 <- rnorm(n, mu1, sd1)
x2 <- rnorm(n, mu2, sd2)
```
2- We can take a look at the data with boxplots
```{r ttest 3 stats test, eval = F}
x_df <- data.frame(Values = c(x1,x2),
                   Group = c(rep('G1',n),rep('G2',n)))
ggplot(x_df) + geom_boxplot(aes(y = Values, color = Group))
```
3- t-test
```{r ttest 4 stats test, eval = F}
t.test(x1, x2)
```
4- The result of the test is influenced by the sample size
```{r ttest 5 stats test, eval = F}
x3 <- rnorm(10000, mu1, sd1)
x4 <- rnorm(10000, mu2, sd2)
t.test(x3, x4)
```
4- We can also do a one-sided test
Check if group2 is greater than group1 (aka the difference 1-2 is smaller)
```{r ttest 6 stats test, eval = F}
t.test(x1, x2, alternative = 'less')
```


#### ANOVA
1- Create mutliple datasets

Our artificial data will have three different means
```{r anova 1 stats test, eval = F}
set.seed(20221011)
m1 <- rnorm(30, mean = 25, sd = 5)
m2 <- rnorm(30, mean = 30, sd = 6)
m3 <- rnorm(30, mean = 22, sd = 7)

ms <- c(m1, m2, m3) # Combine our three measures into one vector

# Our vector of group labels A, B, and C
ls <- as.factor(c(rep("A",30),rep("B",30),rep("C",30)))
```
2- Visualize the data
```{r anova 2 stats test, eval = F}
df_anova <- data.frame(Measure = ms, Group = ls)
ggplot(df_anova) + geom_boxplot(aes(y = Measure, color = Group))
```

3- ANOVA
```{r anova 3 stats test, eval = F}
results_aov <- aov(Measure ~ Group, data = df_anova)
summary(results_aov)
```
We can look at the coefficients to find the difference across groups
```{r anova 4 stats test, eval = F}
results_aov$coefficients
```


#### Hands-on

##### 1- ANOVA vs multiple t-tests
Data PlantGrowth has multiple groups: one control and two treatments

Make an ANOVA and consecutive t-tests between the control and each treatment. 
Do they lead to the same conclusion?
```{r hands on 1 stats test, eval = F}
head(PlantGrowth)
```


##### 2- Multiple testing correction
Take the co2 dataset (note: not CO2), that records the CO2 level for each month across several years. 
We want to test if the CO2 levels from 1959 to 1978 were lower than those between 1979 and 1997 for each month separately. 
Then, we correct for multiple testing and compare the results (hint: see p.adjust). 
Do the conclusions change? 
```{r hands on 2 stats test, eval = F}
co2_df <- data.frame(matrix(co2, 39, 12), row.names = as.character(c(1959:1997)))
colnames(co2_df) <- c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec')
```

#### Hands-on solutions

The solutions for this part are in this [**script**](./exercises/exercises_module3/results/testing_answers.Rmd).

##### 1- ANOVA vs multiple t-tests

Solution hands on 1
```{r hands on 1 solution stats test, eval = F}
results_aov <- aov(weight ~ group, data = PlantGrowth)
summary(results_aov)
results_aov$coefficients
# At least one treatment is different, treatment 1 is lighter than control and
# treatment 2 is heavier

t.test(PlantGrowth$weight[PlantGrowth$group == 'ctrl'], PlantGrowth$weight[PlantGrowth$group == 'trt1'])
t.test(PlantGrowth$weight[PlantGrowth$group == 'ctrl'], PlantGrowth$weight[PlantGrowth$group == 'trt2'])
# Only treatment 2 is significantly different
```

##### 2- Multiple testing correction

Solution hands on 2
```{r hands on 2 solution stats test, eval = F}
p_values <- co2_df %>% apply(2, function(x) return(t.test(x[1:20], x[21:39])$p.value))
sum(p_values < 0.05) # 9 months are different
p_adj <- p.adjust(p_values, method = 'bonferroni')
sum(p_adj < 0.05) # 5 months are different
fdr <- p.adjust(p_values, method = 'fdr')
sum(fdr < 0.05) # 8 months are different
```

