# Lecture 2 - Factor Analysis for Human Spatial Vision 

**Lecturers:** Dr. Alexandre Reynaud

## Lecture

<iframe src="https://docs.google.com/presentation/d/1UdTV3eOuRb9rqvo-8gEnJUZOYJE9GnD4-3_KSmENgFs/preview" width="640" height="480" frameborder="0" allow="autoplay">

</iframe>

```{=html}
<!--<iframe width="640" height="360" src="YOUTUBE EMBED LINK" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen></iframe>
-->
```

## Lab

### Chapter 13: Plotting the Contrast Sensitivity Function 

#### Citing the Guide and the Package {-}

If you have used the **smCSF** package for your research, please cite the paper below:

1. **Min, S. H.**, & Reynaud, A. (2024). Applying Resampling and Visualization Methods in Factor Analysis to Model Human Spatial Vision. ***Investigative Ophthalmology & Visual Science***, 65(1), 17-17.

#### Getting Started {-}

The **smCSF** package will be used in this chapter, so readers should directly download it from my GitHub page using the code below.

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("smin95/smCSF", force = TRUE)
```

Also, we will use the data of achromatic contrast sensitivity from 51 normal observers from this paper (a subset of the entire dataset):

Kim, Y. J., Reynaud, A., Hess, R. F., & Mullen, K. T. (2017). A normative data set for the clinical assessment of achromatic and chromatic contrast sensitivity using a qCSF approach. ***Investigative Ophthalmology & Visual Science***, 58(9), 3628-3636.

Luckily, the data are publicly available. I have reorganized the data so that we could easily use them in the form of a data frame. Let's load the data from online.


```{r, eval=F}
library(tidyverse)
library(smplot2)
library(smCSF)
```

```{r, include=F}
library(tidyverse)
library(smplot2)
library(smCSF)
```

```{r, eval = F}
ACh <- read_csv("./exercises/exercises_lecture2/data_Ach.csv")
```

```{r, echo = F}
ACh <- read_csv("./exercises/exercises_lecture2/data_Ach.csv")
```

```{r}
head(ACh)
```

There are four columns in this data frame:

- First, `Subject` refers to each participant. There are 51 participants total in the .csv file.

- Next, `Repetition` refers to each repetition of the measurement The participants performed two repetitions of the contrast sensitivity measurement.

- `SpatialFreq` refers to spatial frequency of the stimuli that were shown to test the observers' contrast sensitivity. There are a total of 12 spatial frequencies.

- The `Sensitivity` column refers to the linear values of the contrast sensitivity.

#### Brief Introduction

The contrast sensitivity function (CSF) how the observer is sensitive to an image. This image could be large (low spatial frequency) or small (high spatal frequency). To obtain the CSF, the measurement test must measure the threshold. Threshold refers to the difference in stimulus magnitude between the states of presence and absence. If the observer can just barely discriminate between these two states of the visual stimulus, this can be described as the threshold for detection. 

#### Data Visualization

First, let's compute the average, standard errors and standard deviation using the dataset. It is recommended that the reader has read Chapters 5 and 6 already. Understanding of `summarise()`, `filter()`, `mutate()` and `%>%` is assumed.

The code below reorganizes the `ACh` dataset so that instead of containing the data of each subject, it now contains the averaged data across `Repetition` and `SpatialFreq` as well as their standard errors and standard deviations. We use `mean()` to compute the mean, `sm_stdErr()` the standard error, and `sd()` the standard deviation.

```{r}
ACh_avg <- ACh %>%
  group_by(Repetition, SpatialFreq) %>%
  summarise(
    avgSens = mean(Sensitivity),
    stdErrSens = sm_stdErr(Sensitivity),
    stdDevSens = sd(Sensitivity)
  )

ACh_avg
```

Let's create a variable that stores the data for each `Repetition`. Before we do that, however, it is important that we convert the data type of the `Repetition` column from `<dbl>` to `<fct>`, which refers from **double** to **factor**. **Double** refers to continuous data, whereas **factor** refers to categorical data. 

```{r}
ACh_avg$Repetition <- factor(ACh_avg$Repetition) # factor

ACh_avg1 <- ACh_avg %>% filter(Repetition == 1) # repetition 1
ACh_avg2 <- ACh_avg %>% filter(Repetition == 2) # repetition 2
```

Let's now plot the data for each repetition using **ggplot2**. Understanding of Chapters 1-4 is assumed. 

```{r, fig.width = 3.5, fig.height = 3.5, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  geom_point() +
  sm_hgrid() +
  xlab("Spatial frequency (c/deg)") +
  ylab("Averaged sensitivity") +
  ggtitle("Repetition 1 (n=51)")
```

So far, this graph does not show the contrast sensitivity function, which has a rotund shape. Instead, if we connect the points, we get this plot instead.

```{r, fig.width = 3.5, fig.height = 3.5, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  geom_point() +
  geom_line() +
  sm_hgrid() +
  xlab("Spatial frequency (c/deg)") +
  ylab("Averaged sensitivity") +
  ggtitle("Repetition 1 (n=51)")
```

The function is still not a smooth, rotund shape. This is because it is standard to plot the contrast sensitivity function in log10 scales in both x and y axes. The code below shows the plot in the log scale.

```{r, fig.width = 3.5, fig.height = 3.5, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  geom_point() +
  geom_line() +
  sm_hgrid() +
  xlab("Spatial frequency (c/deg)") +
  ylab("Averaged sensitivity") +
  ggtitle("Repetition 1 (n=51)") +
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10")
```

This plot shows the averaged sensitivity for each spatial frequency during the first repetition. However, the contrast sensitivity function is still missing because this plot instead shows the plot that is connected between a pair of the points. To plot the contrast sensitivity function with the data, `sm_CSF()` is required.

```{r, fig.width = 3.5, fig.height = 3.5, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_CSF()
```

Notice that even if the data frame contains data in **linear units**, the plotted data from `sm_CSF()` are in log10 scales. It shows a nice, rotund shape that is typical of the contrast sensitivity function. The function can also plot the linear curve by setting `sm_CSF(logXY = FALSE)`. The default is `sm_CSF(logXY = TRUE)`.

```{r, fig.width = 3.5, fig.height = 3.5, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_CSF(logXY = FALSE)
```

Now let's plot a prettier contrast sensitivity function.

```{r, fig.width = 3.5, fig.height = 3.5, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_CSF(color = sm_color("wine")) +
  geom_point(color = sm_color("darkred"), size = 3) +
  sm_hgrid() +
  xlab("Spatial frequency (c/deg)") +
  ylab("Averaged sensitivity") +
  ggtitle("Repetition 1 (n=51)")
```

We can also plot the data using `ACh_avg2`, which contains data from the second repetition rather than first.

```{r, fig.width = 3.5, fig.height = 3.5, warning = F}
ACh_avg2 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_CSF(color = sm_color("lightgreen")) +
  geom_point(color = sm_color("viridian"), size = 3) +
  sm_hgrid() +
  xlab("Spatial frequency (c/deg)") +
  ylab("Averaged sensitivity") +
  ggtitle("Repetition 2 (n=51)")
```

Notice that `sm_CSF()` is written first because the curve has to appear before the points (`geom_point()`); else, the points will be covered by the CSF curve. This example is shown below.

```{r, fig.width = 3.5, fig.height = 3.5, warning = F}
ACh_avg2 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  geom_point(color = sm_color("viridian"), size = 3) +
  sm_CSF(color = sm_color("lightgreen")) +
  sm_hgrid() +
  xlab("Spatial frequency (c/deg)") +
  ylab("Averaged sensitivity") +
  ggtitle("Repetition 2 (n=51)")
```

The curve appears later in this plot, so it covers some overlapping points. 

**smCSF** also offers a function that fill the area of the CSF curve. This function is `sm_areaCSF()`. The default is set to plot the area in log10 scales, `sm_areaCSF(logXY = TRUE)`.

```{r, fig.width = 3.5, fig.height = 3.5, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_areaCSF() +
  sm_CSF()
```

To plot the linear CSF and its area, `logXY = FALSE` has to be set `FALSE` in both `sm_areaCSF()` and `sm_CSF()`. The functions of `smCSF` have defaults `logXY = TRUE`, therefore, it is recommended that the user uploads a data frame that contains the linear data of spatial frequencies and contrast sensitivities. The functions then will convert them into log scale by default.

```{r, fig.width = 3.5, fig.height = 3.5, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_areaCSF(logXY = FALSE) +
  sm_CSF(logXY = FALSE)
```

We can make the area plot prettier.

```{r, fig.width = 3.5, fig.height = 3.5, warning = F}
ACh_avg2 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_areaCSF(fill = sm_color("lightgreen"), alpha = 0.5) +
  sm_CSF(color = sm_color("viridian")) +
  sm_hgrid() +
  xlab("Spatial frequency (c/deg)") +
  ylab("Averaged sensitivity") +
  ggtitle("Repetition 2 (n=51)") +
  scale_y_log10(limits = c(1, 100)) +
  annotate("text", x = 0.9, y = 1.3, label = "Area under Log CSF")
```

You might get a warning that says `Scale for 'y' is already present. Adding another scale for 'y', which will replace the existing scale`. You can ignore this warning and not be worried about it. This is because we used the function `scale_y_log10()` on top of `sm_CSF()`, which actually calls forth the function `scale_y_continuous()` to convert the plot into log scales. Hence, there are two functions that serve the same purpose here, so the warning message is printed.

The **smCSF** package also offers a function that draws a ribbon shade to denote standard error or standard deviation. This function works exactly the same as `geom_ribbon()`. It requires the uses to specify `ymin` and `ymax`. The example below plots standard deviation using `sm_ribbonCSF()`, where `ymin = avgSens - stdDevSens` and `ymax = avgSens + stdDevSens`. As before, its default is `logXY = TRUE`. Therefore, even if `ACh_avg2` contains data in linear units, the ribbon shade that is generated by `sm_ribbonCSF()` will be in log unit. 

```{r, fig.width = 3.5, fig.height = 3.5, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_ribbonCSF(aes(
    ymin = avgSens - stdDevSens,
    ymax = avgSens + stdDevSens
  )) +
  sm_CSF() +
  scale_y_log10(limits = c(1, 100))
```

Let's make the plot prettier using **smplot**.

```{r, fig.width = 3.5, fig.height = 3.5, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_ribbonCSF(
    aes(
      ymin = avgSens - stdDevSens,
      ymax = avgSens + stdDevSens
    ),
    fill = sm_color("wine"), alpha = 0.4
  ) +
  sm_CSF(color = sm_color("darkred")) +
  xlab("Spatial frequency (c/deg)") +
  ylab("Averaged sensitivity") +
  ggtitle("Repetition 1 (n=51)") +
  sm_hgrid() +
  scale_y_log10(limits = c(1, 100))
```

You can also shade standard error instead, but the ribbon is barely visible.

```{r, fig.width = 3.5, fig.height = 3.5, warning = F}
ACh_avg2 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_ribbonCSF(aes(
    ymin = avgSens - stdErrSens,
    ymax = avgSens + stdErrSens
  )) +
  sm_CSF() +
  scale_y_log10(limits = c(1, 100))
```


As practice, let's plot the averaged data across all repetitions.

```{r}
ACh_avg_final <- ACh %>%
  group_by(SpatialFreq) %>%
  summarise(
    avgSens = mean(Sensitivity),
    stdErrSens = sm_stdErr(Sensitivity),
    stdDevSens = sd(Sensitivity)
  )

head(ACh_avg_final)
```

Here is the new data frame `ACh_avg_final`.

```{r, fig.width = 3.5, fig.height = 3.5, warning = F}
ACh_avg_final %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_ribbonCSF(
    aes(
      ymin = avgSens - stdDevSens,
      ymax = avgSens + stdDevSens
    ),
    fill = sm_color("skyblue"), alpha = 0.25
  ) +
  sm_CSF(
    color = sm_color("blue"),
    alpha = 0.4
  ) +
  xlab("Spatial frequency (c/deg)") +
  ylab("Averaged sensitivity") +
  ggtitle("Contrast Sensitivity (n=51)") +
  sm_hgrid() +
  geom_point(color = sm_color("blue"), size = 2.5) +
  scale_y_log10(limits = c(1, 100))
```


The standard deviation seems to be still broad.

#### Facetting the Contrast Sensitivity Functions

To facet means to split the plot into multiple plots based on a categorical variable (i.e., **factor**). We will facet the contrast sensitivity function plot above (the blue curve) into two panels based on two repetitions (red and green below).

```{r, fig.width = 6.5, fig.height = 3.5, warning = F, echo = F}
ACh_avg %>%
  ggplot(aes(
    x = SpatialFreq, y = avgSens,
    group = Repetition, color = Repetition,
    fill = Repetition
  )) +
  sm_ribbonCSF(aes(
    ymin = avgSens - stdDevSens,
    ymax = avgSens + stdDevSens
  ), alpha = 0.4) +
  sm_CSF(alpha = 0.4) +
  sm_hgrid() +
  sm_facet_header(font_size = 11) +
  geom_point(size = 2.5) +
  scale_fill_manual(values = sm_color("wine", "lightgreen")) +
  scale_color_manual(values = sm_color("darkred", "viridian")) +
  xlab("Spatial frequency (c/deg)") +
  ylab("Averaged sensitivity") +
  facet_wrap(~Repetition,
    labeller = as_labeller(c(
      "1" = "Repetition 1",
      "2" = "Repetition 2"
    ))
  ) +
  scale_y_log10(limits = c(1, 100))
```

To facet the plot, you need to have a data frame that has data with a column that specifies the categorical variable. For instance, let's look at the data frame `ACh_avg`.

```{r}
ACh_avg
```

We see that the first column of `Ach_avg` has a column named `Repetition` which stores whether the given data are from the first or second repetition. To see all the unique levels of `Repetition`, we can use the function `unique()`.

```{r}
unique(ACh_avg$Repetition)
```

To facet a plot using ggplot2, the function `facet_wrap()` has to be used.

```{r, fig.width = 6.5, fig.height = 3.5, warning = F}
ACh_avg %>%
  ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_CSF() +
  facet_wrap(~Repetition)
```

The argument inside `facet_wrap` has to include the tilde (`~`). The right side the tilde (`~`) should be the name of the categorical variable by which you wish to facet the plot. If the variable however happens to be continuous, `facet_wrap()` would not work properly.

```{r, fig.width = 6.5, fig.height = 3.5, warning = F}
ACh_avg %>%
  ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_CSF() +
  facet_wrap(~stdErrSens)
```

However, I do not often use `facet_wrap()` because the plot often ends up not being pretty. Let's try to make it prettier.

```{r, fig.width = 6.5, fig.height = 3.5, warning = F}
ACh_avg %>%
  ggplot(aes(
    x = SpatialFreq, y = avgSens,
    group = Repetition, color = Repetition,
    fill = Repetition
  )) +
  sm_ribbonCSF(aes(
    ymin = avgSens - stdDevSens,
    ymax = avgSens + stdDevSens
  ), alpha = 0.4) +
  sm_CSF(alpha = 0.4) +
  sm_hgrid() +
  geom_point(size = 2.5) +
  scale_fill_manual(values = sm_color("wine", "lightgreen")) +
  scale_color_manual(values = sm_color("darkred", "viridian")) +
  xlab("Spatial frequency (c/deg)") +
  ylab("Averaged sensitivity") +
  facet_wrap(~Repetition) +
  scale_y_log10(limits = c(1, 100))
```

Personally, I do not like the grey-filled header of each panel. Since I personally will find myself repeatedly bypassing the default grey color into a color of my choice, I have created the function myself `sm_facet_header()` as part of the **smCSF** package.

```{r, fig.width = 6.5, fig.height = 3.5, warning = F}
ACh_avg %>%
  ggplot(aes(
    x = SpatialFreq, y = avgSens,
    group = Repetition, color = Repetition,
    fill = Repetition
  )) +
  sm_ribbonCSF(aes(
    ymin = avgSens - stdDevSens,
    ymax = avgSens + stdDevSens
  ), alpha = 0.4) +
  sm_CSF(alpha = 0.4) +
  sm_hgrid() +
  geom_point(size = 2.5) +
  scale_fill_manual(values = sm_color("wine", "lightgreen")) +
  scale_color_manual(values = sm_color("darkred", "viridian")) +
  xlab("Spatial frequency (c/deg)") +
  ylab("Averaged sensitivity") +
  facet_wrap(~Repetition) +
  scale_y_log10(limits = c(1, 100)) +
  sm_facet_header()
```

Now the title of each plot has been bolded and the filled color is white. I personally think that this plot is a lot cleaner than the default one.

Next, I wish to change the titles of each panel because `1` and `2` are not descriptive. To do so, we need to use the argument `labeller` within `facet_wrap()` to specify the title of each panel.

```{r, fig.width = 6.5, fig.height = 3.5, warning = F}
ACh_avg %>%
  ggplot(aes(
    x = SpatialFreq, y = avgSens,
    group = Repetition, color = Repetition,
    fill = Repetition
  )) +
  sm_ribbonCSF(aes(
    ymin = avgSens - stdDevSens,
    ymax = avgSens + stdDevSens
  ), alpha = 0.4) +
  sm_CSF(alpha = 0.4) +
  sm_hgrid() +
  geom_point(size = 2.5) +
  scale_fill_manual(values = sm_color("wine", "lightgreen")) +
  scale_color_manual(values = sm_color("darkred", "viridian")) +
  xlab("Spatial frequency (c/deg)") +
  ylab("Averaged sensitivity") +
  facet_wrap(~Repetition,
    labeller = as_labeller(c(
      "1" = "Repetition 1",
      "2" = "Repetition 2"
    ))
  ) +
  scale_y_log10(limits = c(1, 100)) +
  sm_facet_header(font_size = 11)
```

We use the function `as_labeller()` as part of the argument `labeller()` inside `facet_wrap()` to specify that factor level `1` should be shown as `Repetition 1` and `2` as `Repetition 2`.

We can also change the number of rows and columns using `facet_wrap()`. Lets make a data frame that has 4 repetitions instead of 2. 

```{r}
set.seed(1)
spatFreqNum <- length(unique(ACh_avg$SpatialFreq))
ACh_avg_4rep <- rbind(ACh_avg, ACh_avg)
ACh_avg_4rep$Repetition <- rep(1:4, each = spatFreqNum)
ACh_avg_4rep[25:nrow(ACh_avg_4rep), 3:5] <- ACh_avg_4rep[25:nrow(ACh_avg_4rep), 3:5] + round(rnorm(24), 1)
ACh_avg_4rep$Repetition <- factor(ACh_avg_4rep$Repetition)
```

We can specify the number of columns (ex. `ncol = 2`) and rows (ex. `nrow = 2`) inside the function `facet_wrap()`, as shown in the example below.

```{r, fig.width = 6.5, fig.height = 6.5, warning = F}
ACh_avg_4rep %>%
  ggplot(aes(
    x = SpatialFreq, y = avgSens,
    group = Repetition, color = Repetition,
    fill = Repetition,
    alpha = Repetition
  )) +
  sm_ribbonCSF(aes(
    ymin = avgSens - stdDevSens,
    ymax = avgSens + stdDevSens
  )) +
  sm_CSF(alpha = 0.4) +
  sm_hgrid() +
  geom_point(alpha = 1, size = 2.5) +
  scale_fill_manual(values = sm_color("wine", "lightgreen", "yelloworange", "crimson")) +
  scale_color_manual(values = sm_color("darkred", "viridian", "orange", "red")) +
  scale_alpha_manual(values = c(0.4, 0.4, 0.3, 0.1)) +
  xlab("Spatial frequency (c/deg)") +
  ylab("Averaged sensitivity") +
  facet_wrap(~Repetition,
    nrow = 2, ncol = 2,
    labeller = as_labeller(c(
      "1" = "Repetition 1",
      "2" = "Repetition 2",
      "3" = "Repetition 3",
      "4" = "Repetition 4"
    ))
  ) +
  scale_y_log10(limits = c(1, 100)) +
  sm_facet_header(font_size = 11)
```

Notice that when `facet_wrap()` is used, there is no need to create variable for each plot like we have done in Chapter 5 and use `sm_common_axis()`. The preference can vary depending on the user. Rather than using `facet_wrap()`, however, I still prefer to store each plot separately so that I can have freedom to decide where to place each panel wherever I want using `sm_common_axis()`.

Also, make sure that when you combine the plots using `facet_wrap()` you set the size of the figure properly within the `save_plot()` function or else the figure will appear to be too crowded. You can use `base_height` and `base_width` arguments within `facet_wrap()`.

Another issue with plotting each contrast sensitivity function per panel is that it is visually difficult to compare whether the functions between the panels are different. One way to overcome this problem is to plot all the plots together. In this case, legend is necessary, which can be generated using `sm_hgrid(legends = TRUE)`. Let's use the data frame `ACh_avg` which has two levels in the `Repetition` column.

```{r, fig.width = 3.5, fig.height = 3.5, warning = F}
ACh_avg %>%
  ggplot(aes(
    x = SpatialFreq, y = avgSens,
    group = Repetition, color = Repetition,
    fill = Repetition
  )) +
  sm_ribbonCSF(aes(
    ymin = avgSens - stdDevSens,
    ymax = avgSens + stdDevSens
  ), alpha = 0.2) +
  sm_CSF(alpha = 0.4) +
  sm_hgrid(legends = T) +
  scale_fill_manual(values = sm_color("wine", "lightgreen")) +
  scale_color_manual(values = sm_color("darkred", "viridian")) +
  xlab("Spatial frequency (c/deg)") +
  ylab("Averaged sensitivity") +
  scale_y_log10(limits = c(1, 100)) +
  theme(legend.position = c(0.8, 0.2)) +
  ggtitle("Achromatic CSF (n=51)")
```

We see that these two curves are nearly identical, except that the green curve is uniformly higher than the red curve. In the next chapter, we will discuss about some techniques on how to analyze contrast sensitivity data.


### Chapter 16: Factor Analysis and CSF 

#### Citing the Guide and the Package {-}

If you have used the **smCSF** package for your research, please cite the paper below:

1. **Min, S. H.**, & Reynaud, A. (2024). Applying Resampling and Visualization Methods in Factor Analysis to Model Human Spatial Vision. ***Investigative Ophthalmology & Visual Science***, 65(1), 17-17.

#### Getting Started {-}

```{r, message=F, warning=F}
library(tidyverse)
library(cowplot)
library(smplot2)
library(smCSF)
library(psych)
```

In this chapter, we will use a public dataset of achromatic contrast sensitivity from 51 normally sighted observers.

```{r, message=F, warning=F}
df <- read_csv("./exercises/exercises_lecture2/data_ACh.csv") %>%
  group_by(Subject, SpatialFreq) %>%
  summarise(
    Sensitivity = mean(Sensitivity),
    Repetition = "avg"
  )
```

Factor analysis is a technique to identify latent factors that produce local patterns/correlations in a high-dimensional dataset. In other words, it aims to reduce the number of dimensions in the data.

In this dataset, there are 12 unique spatial frequencies. The sensitivity data were collected at these spatial frequencies to produce the contrast sensitivity function, which has a peak and has an asymmetrical shape. There are 12 dimensions total.

Can the curve be summarised using less than 12 dimensions? This is the purpose of factor analysis.

One way to extract an adequate number of factors/dimensions is to draw a scree plot. 

#### Eigenvalues

The scree plot has a y-axis of the **eigenvalue**, which refers the amount of variance of the data from the factor of interest. Essentially, the higher the eigenvalue, the more important the factor is in the dataset.

Some methods for factor analysis determine the number of factors in a potential factor model based on eigenvalues. The Guttman Rule (i.e. K1) extracts all factors with eigenvalues larger than 1.

Another method that relies on eigenvalues to determine the number of factors is parallel analysis.

#### Parallel analysis

Parallel analysis by Horn is a statistical method to determine the number of factors in an exploratory factor model. It uses Monte Carlo simulation to generate a dataset of random numbers. This random dataset has the same size of the real dataset. Eigenvalues are obtained from the random dataset.

We will use `psych` package's function `fa.parallel` to perform parallel analysis.

```{r, warning = F, message = F}
df_mat <- read_csv("https://www.smin95.com/data_ACh_mat.csv") # ACh data in matrix form
df_mat1 <- as.matrix(df_mat[, -1])
```

After we have extracted the data that are in matrix form (not data frame), which is required to properly use functions of the `psych` package, we can then perform parallel analysis.

```{r, fig.width=4.2, fig.height=4.2, warning = F, message = F}
fa.parallel(df_mat1, fa = "fa")
```

Based on the parallel analysis, we see that there are two factors with eigenvalues larger than those from the random data. Therefore, it supports a two-factor model.

#### Non-parametric bootstrapping 

The Guttman rule can suggest different numbers of factors depending on the sample. For example, in one dataset, factor A can have an eigenvalue larger than 1 but in another dataset from another population factor B can have an eigenvalue lower than 1. Therefore the number of factors that is extracted can vary depending on the local sample. This creates variability in results. This issue can also be observed in parallel analysis.

Therefore, it might be better to include a range of uncertainty of eigenvalues so that one can better be informed about the possible range of eigenvalues in a population. This can be done by plotting 95% confidence interval of eigenvalues. 

In this section, we will perform non-parametric bootstrapping to compute the confidence intervals of the eigenvalues. However, it turns out, if we resample the contrast sensitivity data directly, it removes the local correlation and therefore the influence of latent factors on the dataset, thereby erasing covariance from its existence. 

Here is the original data set's covariance of achromatic sensitivity.

```{r}
cov(df_mat1)
```

Covariance is high. If the dataset's covariance is high, then it is suitable for factor analysis, as demonstrated by Kaiser, Meyer and Olkin test.

```{r}
KMO(df_mat1)
```

The test reveals that covariance is adequately high (test score > 0.5) at most spatial frequencies, suggesting that factor analysis is suitable.

Now I demonstrate what happens to covariance after resampling with replacement.

```{r}
set.seed(2023)

nObs <- 51
nSF <- ncol(df_mat1) # 12 SFs total
resampled_ACh <- matrix(data = NA, nrow = nObs, ncol = nSF)

for (iSF in 1:nSF) {
  resampled_ACh[, iSF] <- sample(df_mat1[, iSF], nObs, replace = T)
}
```

Here is the re-sampled data set. We can also plot the re-sampled data against the experimental data using base `R` (not `ggplot2`) since we are using matrices to store data. Plotting in base `R` can be more tedious but it can be quite fun as you can control every aspect of the plot.

```{r, fig.width=3.6, fig.height=3.6}
avgs_expt <- avgs_resampled <- vector("double", nSF)

for (iSF in 1:nSF) {
  avgs_resampled[[iSF]] <- mean(resampled_ACh[, iSF])
  avgs_expt[[iSF]] <- mean(df_mat1[, iSF])
}

par(mfrow = c(1, 1), mar = c(4, 4.1, 1.2, 1.2))

plot(unique(df$SpatialFreq), log(avgs_resampled),
  log = "x",
  ylim = c(0.1, max(log(c(avgs_resampled, avgs_expt))) + 1),
  xlab = "Spatial frequency",
  ylab = "Contrast sensitivity", col = sm_color("blue"),
  pch = 15, cex = 1.4
)

lines(unique(df$SpatialFreq), log(avgs_resampled),
  col = sm_color("blue"), cex = 2
)

points(unique(df$SpatialFreq) + unique(df$SpatialFreq) * 0.1,
  log(avgs_expt),
  pch = 16, col = sm_color("red"), cex = 1.4
)

lines(unique(df$SpatialFreq) + unique(df$SpatialFreq) * 0.1,
  log(avgs_expt),
  col = sm_color("red"), cex = 2
)

legend(0.3, 1.5,
  legend = c("Resampled data", "Real data"),
  col = sm_color("blue", "red"), pch = 15:16, pt.cex = 1.8,
  bty = "n"
)
```

The two data sets look very similar, however, we will see that \only one of them (real data) can be used for factor analysis because it properly retains covariance. Therefore, it seems that factor analysis captures the *invisible* pattern of the data.

The covariance of the re-sampled data is shown below.

```{r}
cov(resampled_ACh)
```

We now see that the re-sampled data set has a much lower covariance in general. Since covariance is too low, it is not suitable to perform factor analysis.

```{r}
KMO(resampled_ACh)
```

The values from the KMO test are lower than 0.5 at most spatial frequencies, suggesting the data set is not suitable for factor analysis. 

We can still try parallel analysis and see what happens.

```{r, fig.width=4.2, fig.height=4.2, warning = F, message = F}
fa.parallel(resampled_ACh, fa = "fa")
```

There is no identified factor because there is no covariance. However, re-sampling is required to compute 95% confidence interval.

A solution is to use a model that captures the contrast sensitivity function, compute parameters that retain the overall characteristic of the curve for each subject, and then re-sample the parameters with replacement. Then using the re-sampled parameters, one can refit the sensitivity data. In this example, we will re-sample for 100 times (`nSim <- 100`).

Calculation of the parameters from a data frame that contains contrast sensitivity data can be perfomed with `sm_params_list()`.

```{r}
param_res <- sm_params_list(
  subjects = "Subject",
  conditions = "Repetition",
  x = "SpatialFreq",
  values = "Sensitivity",
  data = df
)
```

After we compute the four parameters of the parabola model of CSF for each subject, we resample them with replacement. Note that the sample size with resampled data should be identical to the original sample size (`nObs <- 51`).


```{r}
nObs <- 51
nSim <- 100 # 100 simulations
```

Then, we can perform non-parametric bootstrapping and obtain 95% confidence interval of the eigenvalues using `sm_np_boot()`. It produces a data frame with the confidence intervals from re-sampling and mean eigenvalue from the multiple rounds of re-sampling. This is stored on `boot_res`.

```{r,  eval = F}
boot_res <- sm_np_boot(param_res, n = nObs, nSim = nSim)
boot_res
```

```{r, echo = F, message=F, warning=F}
boot_res <- read_csv("./exercises/exercises_lecture2/boot_res.csv")
boot_res
```

By using `sm_plot_boot()`, we can then draw the scree plot with the eigenvalues and their 95% confidence intervals. It draws it by using `ggplot2`, so you can customise the aesthetics by adding (`+`) ggplot2 functions as I have done here with `ylab()`.

```{r, fig.width = 3.8, fig.height=3.2}
sm_plot_boot(boot_res, shapes = 16) +
  ylab("Eigenvalue") -> eigen1

eigen1
```

We can then extract the number of factors where the 95% confidence intervals of the eigenvalues from the observed data do not overlap with those from randomly generated data from parallel analysis. In this example, we can develop an exploratory model with two factors.

#### Factor Analysis

Now, we perform factor analysis with two factors. In this example, we perform varimax rotation to increase the interpretability of the results from factor analysis.

```{r, warning=F, message=F}
nFac <- 2
fa_res <- fa(df_mat1, nFac, fm = "minres", rotate = "varimax")
print(fa_res$loadings, cutoff = 0.5)
```

It seems that two factors can describe enough of the achromatic data. The proportion of variance from these two factors are 87.7%. However, there needs some more work to do to confirm whether a two-factor model adequately captures the covariance of the achromatic data.
