# Practical 5

## Lecture

<iframe 
  src="https://docs.google.com/presentation/d/1pfe0eEzW3eC-ddswPPaX1UMDAOf3JH-Z/preview" 
  width="640" 
  height="480" 
  frameborder="0" 
  allow="autoplay">
</iframe>

<!--<iframe width="640" height="360" src="YOUTUBE EMBED LINK" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen></iframe>
-->

## Lab

Materials from this section can be downloaded [from here](https://github.com/aosakwe/CBW-INR-RMarkdown-Quarto).


### Example Script

You can download the [example_script.R here](./exercises/exercises_module5/example_script.R).

```{r}

library(ggplot2)
library(dplyr)

#Script for modelling iris species based on sepal characteristics
## Iris Dataset Modelling
#The iris dataset is composed of 50 flowers from three different species. 
#Measurements include width and length (centimeters) for petals and sepals of each flower.

#Filtering out versicolor to run a binary example
#focus is on viridis and seratosa species
mini_iris <- iris %>%
  filter(Species != 'versicolor')



### Scatter Plot of Iris dataset with Sepal Length and Width as axes
### Points are colored by their species
ggplot(mini_iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 3, alpha = 0.7) +  # Add points with some transparency
  theme_minimal() +  # Use a clean theme
  labs(
    title = "Iris Dataset: Sepal Dimensions by Species",
    x = "Sepal Length",
    y = "Sepal Width"
  )


#Objective
#Goal of this analysis is to identify features capable of distinguishing flower species with high accuracy. We achieve this task using the **logistic regression model**, focusing on 1 vs. all comparisons.


#Training logistic regression model
iris.fit <- glm(Species ~ Sepal.Length +  Sepal.Width,
                family = 'binomial',
                data = mini_iris)


#Calculating class predictions
mini_iris$Prob <- predict.glm(iris.fit,
                              mini_iris[,c('Sepal.Length','Sepal.Width')],
                              type = 'response')

#Assigning class labels
mini_iris$Pred <- ifelse(mini_iris$Prob >= 0.5,"virginica", "setosa")

#Comparing to ground truth labels
mini_iris$Result <- mini_iris$Species == mini_iris$Pred


## Visualizing Sepal dimensions by species
ggplot(mini_iris, aes(x = Sepal.Length, y = Sepal.Width, color = Pred)) +
  geom_point(size = 3, alpha = 0.7) +  # Add points with some transparency
  theme_minimal() +  # Use a clean theme
  labs(
    title = "Iris Dataset: Training Classification Performance",
    x = "Sepal Length",
    y = "Sepal Width"
  )


## Repeating analysis between Viridis and versicolor
mini_iris <- iris %>%
  filter(Species != 'setosa')
ggplot(mini_iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point(size = 3, alpha = 0.7) +  # Add points with some transparency
  theme_minimal() +  # Use a clean theme
  labs(
    title = "Iris Dataset: Dimensions by Species",
    x = "Petal Length",
    y = "Petal Width"
  )

iris.fit <- glm(Species ~ Sepal.Length + Sepal.Width,
                family = 'binomial',
                data = mini_iris)

iris_pred <- predict.glm(iris.fit,
                         mini_iris[,c('Sepal.Length','Sepal.Width')],
                         type = 'response')
table(mini_iris$Species == ifelse(iris_pred >= 0.5, 'virginica','versicolor'))

```


### Example Notebook

You can download the [Example_Notebook.Rmd here](./exercises/exercises_module5/Example_Notebook.Rmd).


```{r}
library(tidyverse)
```

#### Iris Dataset Modelling

The iris dataset is composed of 50 flowers from three different species. Measurements include width and length (centimeters) for petals and sepals of each flower.

```{r}
head(iris)
```

##### Feature visualization

```{r}
mini_iris <- iris %>%
  filter(Species != 'versicolor')

ggplot(mini_iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 3, alpha = 0.7) +  # Add points with some transparency
  theme_minimal() +  # Use a clean theme
  labs(
    title = "Iris Dataset: Sepal Dimensions by Species",
    x = "Sepal Length",
    y = "Sepal Width"
  )

```

#### Objective

Goal of this analysis is to identify features capable of distinguishing flower species with high accuracy. We achieve this task using the **logistic regression model**, focusing on 1 vs. all comparisons.

##### Logistic Regression

From a probabilistic perspective, the logistic regression follows a Bernoulli distribution

$$
y_i \sim Bernouilli(\pi_i)
$$

Where $y_i$ is the classification for observation $i$. $\pi_i$ represents the Bernoulli parameter for observation $i$ and is calculated as

$$
\pi_i = \sigma(x\beta) = P(y_i = 1| x)=  \frac{1}{1 + \exp(-x\beta)}
$$

Logistic regression extends the linear model by using the sigmoid/logistic function $\sigma(x\beta)$ which allows it to bound its outputs between 0 and 1.

#### Model Fitting

```{r}

## Question: can you try using Petal length and width instead?
##Hint: how are we currently specifying what features are used in the model?
iris.fit <- glm(Species ~ Sepal.Length +  Sepal.Width,
                family = 'binomial',
                data = mini_iris)


mini_iris$Prob <- predict.glm(iris.fit,
                              mini_iris,
                              type = 'response')

##Question: Can you change the two lines below into a pipe operation?
##Hint: you will need the mutate() function and the %>% operator
mini_iris$Pred <- ifelse(mini_iris$Prob >= 0.5,"virginica", "setosa")
mini_iris$Result <- mini_iris$Species == mini_iris$Pred


table(mini_iris$Result)
```

#### Model Visualization

```{r}
##Question: Can you add an additional plot that shows the TRUE species labels?
#Hint: what was the name of the column storing the labels? What argument is being used to visualize the Predicted labels?
ggplot(mini_iris, aes(x = Sepal.Length, y = Sepal.Width, color = Pred)) +
  geom_point(size = 3, alpha = 0.7) +  # Add points with some transparency
  theme_minimal() +  # Use a clean theme
  labs(
    title = "Iris Dataset: Training Classification Predictions",
    x = "Sepal Length",
    y = "Sepal Width"
  )

## Running a linear regression to compare to logistic regression
mini_iris$Num_Species <- ifelse(mini_iris$Species == 'setosa', 1,2)
iris.lm <- lm(Num_Species ~ Sepal.Length +  Sepal.Width,
              data = mini_iris)

mini_iris$Num_pred <- predict.lm(iris.lm,mini_iris)

summary(mini_iris$Num_pred)

summary(mini_iris$Prob)


```

##### Feature selection and additional modelling

Exploring optimal features to compare viridis and versicolor species.

```{r}

mini_iris <- iris %>%
  filter(Species != 'setosa')

##Question:can you change the plot to show Petal length and width?
##hint: aes(x = ?, y = ?)
ggplot(mini_iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 3, alpha = 0.7) +  # Add points with some transparency
  theme_minimal() +  # Use a clean theme
  labs(
    title = "Iris Dataset: Dimensions by Species",
    x = "Sepal Length",
    y = "Sepal Width"
  )

```

##### Evaluation

Evaluation of the trained logistic regression model demonstrates satisfactory classification performance.

N.B: current estimates are based on **training data**. Must adapt analysis to be based on held-out samples for future updates.

```{r}


iris.fit <- glm(Species ~ Sepal.Length + Sepal.Width,
                family = 'binomial',
                data = mini_iris)


iris_pred <- predict.glm(iris.fit,
                         mini_iris,
                         type = 'response')


## Question: is the below function hard to read? How could we fix it?
##Hint: pipe operator
as.data.frame(table(mini_iris$Species == ifelse(iris_pred >= 0.5, 'virginica','versicolor')))
```

