---
title: "Logistic Regression"
author: 
- name: "Audrey Baguette"
  affiliation: Workshop Lead
- name: "Alex Diaz-Papkovich"
  affiliation: Original Material
- name: "Adrien Osakwe"
  affiliation: Original Material
date: "`r format(Sys.Date(), '%B %d, %Y')`"
---

# Loading libraries
```{r Loading libraries}
library(tidyverse)
library(nnet)
```


# Predicting species with a linear model
We remove the versicolor species to only have two species
```{r lm 1}
iris_subset <- iris %>% filter(Species != 'versicolor')
```
Predict the species based on the petal dimensions
```{r lm 2}
iris_subset$train_test <- sample(c('train', 'test'), nrow(iris_subset), replace = TRUE, prob = c(0.8, 0.2))
iris_model <- lm(Species ~ Petal.Length + Petal.Width, data = iris_subset[iris_subset$train_test == 'train',])
predictions <- predict.lm(iris_model, iris_subset[iris_subset$train_test == 'test',], type = 'response')
```
Convert the predictions to species and compare to the truth
```{r lm 3}
table(round(predictions), iris_subset[iris_subset$train_test == 'test', 'Species'])
```


# Predicting 2 species with a log model
```{r log 2 categories}
iris_log_model <- glm(Species ~ Sepal.Length +  Sepal.Width,
                  family = 'binomial',
                  data = iris_subset[iris_subset$train_test == 'train',])

predictions_log <- predict.glm(iris_log_model,
                               iris_subset[iris_subset$train_test == 'test',],
                              type = 'response')

predictions_log <- ifelse(predictions_log >= 0.5,"virginica", "setosa")

table(predictions_log, iris_subset[iris_subset$train_test == 'test', 'Species'])
```


# Predicting 3 species with a log model
```{r log 3 categories 1}
iris$train_test <- sample(c('train', 'test'), nrow(iris), replace = TRUE, prob = c(0.8, 0.2))
iris_train <- iris %>% filter(train_test == 'train')
iris_test <- iris %>% filter(train_test == 'test')

iris_multinomial <- multinom(Species ~ Sepal.Length +  Sepal.Width,
                      data = iris_train)

predictions_multi <- predict(iris_multinomial,
                               iris_test,
                               'prob')
```
Convert the probabilities to species with a softmax
```{r log 3 categories 2}
species_predictions <- predictions_multi %>%
  apply(1, function(x) return(c('setosa', 'versicolor', 'virginica')[which.max(x)]))

table(species_predictions, iris_test$Species)
```


# Hands-on
From the diamonds model, predict the cut

First divide the data in training and testing datasets
```{r hands on}
diamonds$train_test <- sample(c('train', 'test'), nrow(diamonds), replace = TRUE, prob = c(0.8, 0.2))
diamonds_train <- diamonds %>% filter(train_test == 'train')
diamonds_test <- diamonds %>% filter(train_test == 'test')
```
Solution
```{r hands on ex}

```



#