---
title: "Data Wrangling"
author: 
- name: "Audrey Baguette"
  affiliation: Workshop Lead
- name: "Larisa M. Soto"
  affiliation: Original Material
- name: "Adrien Osakwe"
  affiliation: Original Material
date: "`r format(Sys.Date(), '%B %d, %Y')`"
---

# Loading libraries
```{r Loading libraries}
library(dplyr)
library(tidyr)
```


# Selected dplyr functions

## 1- Select
Select permits to extract, and potentially rename, one or more columns of a data frame.
```{r select doc}
help(select)
```
### Examples
Select the column "Species" from data frame iris
```{r select 1}
head(select(iris, Species))
```
Select the columns "Sepal.Length" and "Species" from data frame iris
```{r select 2}
head(select(iris, Sepal.Length, Species))
```
Select the columns "Sepal.Length" and "Species" from data frame iris AND rename "Sepal.Length" as "Sepal"
```{r select 3}
head(select(iris, Sepal = Sepal.Length, Species))
```


## 2- Filter
Extract rows of a data frame, based on some conditions they need to meet.
```{r filter doc}
help(filter)
```
### Examples
Select rows of Species "versicolor"
```{r filter 1}
filter(iris, Species == 'versicolor')
```
Select rows of Species "versicolor" AND Sepal.Length is smaller than 6
```{r filter 2}
filter(iris, Species == 'versicolor' & Sepal.Length < 6)
```


## 3- Group by
Split a data frame in parts, based on the values contained in a column.
```{r groupby doc}
help(group_by)
```
### Example
Group by Species
```{r groupby 1}
group_by(iris, Species)
```


## 4- Summarize
Group by the values in a column, then produce a specified summary statistic.
```{r summarize doc}
help(summarize)
```
### Examples
Get the mean Petal.Length
```{r summarize 1}
summarize(iris, mean(Petal.Length))
```
Get the mean Petal.Length, Petal.Length standard deviation and mean Sepal.Length AND rename them
```{r summarize 2}
summarize(iris,
          Mean_petal_length = mean(Petal.Length),
          SD_petal_length = sd(Petal.Length),
          Mean_sepal_length = mean(Sepal.Length))
```


## 5- Count
Count the number of individuals (rows) in a given category.
```{r count doc}
help(count)
```
### Examples
Count the number of individual in each species
```{r count 1}
count(iris, Species)
```
Count the number of individuals with Sepal.Length above 6
```{r count 2}
count(iris, Sepal.Length > 6)
```
Count the number of individuals with Sepal.Length above 6, group by Species AND rename the column
```{r count 3}
count(iris, Sepal.Length > 6, Species, name = 'sepal_length_bigger_6')
```


## 6- Mutate
Transform, combine and do operations on columns
```{r mutate doc}
help(mutate)
```
### Examples
Compute the Petal area (assuming diamond shapes)
```{r mutate 1}
head(mutate(iris, Petal_surface = Petal.Length * Petal.Width / 2))
```
Compute the sums of lengths and widths
```{r mutate 2}
head(mutate(iris, Length_sum = Petal.Length + Sepal.Length, Width_sum = Petal.Width + Sepal.Width))
```


Here is a link to a cheatsheet, that contains small descriptions of many dplyr functions:
https://github.com/rstudio/cheatsheets/blob/main/data-transformation.pdf



# Piping functions

Say that we want to group the iris data frame by species, then create summary statistics. We can do it...
## ...Witout piping
```{r no piping}
grouped_by_species <- group_by(iris, Species)
mean_stats <- summarize(grouped_by_species,
                        mean(Petal.Length), 
                        mean(Petal.Width), 
                        mean(Sepal.Length), 
                        mean(Sepal.Width))
mean_stats
```

## ...With piping
```{r with piping}
iris %>%
  group_by(Species) %>%
  summarize(mean(Petal.Length), mean(Petal.Width), mean(Sepal.Length), mean(Sepal.Width))
```
The piping operator `%>%` permits to put multiple functions one after the other, where the output of the function of the left of the pipe operator is used as first argument of the function on the right side.

## Hands on 1: Conversion of script
Convert the following script to its piped version:
```{r hands on 1}
grouped_cars <- group_by(mtcars, vs, hp >= 100) # group by vs (engine shape) and hp (horsepower)
ratio_mpg_wt <- mutate(grouped_cars,ratio = mpg/wt) # add a column with the ratio of mpg (miles per gallon) and wt (weight)
summarize(ratio_mpg_wt,
          Mean_weight = mean(wt),
          Mean_mpg = mean(mpg),
          Mean_ratio = mean(ratio)) # summarize statistics
```
Solution
```{r hands on 1 ex}

```

## Hands on 2:
Count the number of cars with (1) a sum of gears (gear) and carburators (carb) >= 6, (2) a number of cylinders (cyl) < 8 and (3) stratified by engine shape (vs)

Solution
```{r hands on 2 ex}

```

## Hands on 3: Find the errors
We want to get the number of flowers with a sepal area (assuming area = width * length / 2) > 10, stratified by species
```{r hands on 3.1}
iris %>%
  mutate(area = Sepal.Length * Sepal.Width / 2) %>%
  group_by(Species) %>%
  summarize(sum(area))
```
Solution
```{r hands on 3.1 ex}

```

We want summary statistics by brand
```{r hands on 3.2}
#mtcars %>%
#  mutate(brand = sub(' .*', '', row.names(mtcars))) %>%
#  group_by(brand) %>%
#  summarize(sum(brand), mean(gear), mean(carb))

# ERROR
```
Solution
```{r hands on 3.2 ex}

```



# tidyr pivot functions
The pivot functions (`pivot_longer` and `pivot_wider`) can be used to transform a data frame, such as to convert columns into row values and vice-versa. For example, if we have a data in wide format...
```{r pivot 1}
billboard_subset <- billboard[1:15, 1:11]
print(billboard_subset)
```
...we can convert it to long format...
```{r pivot 2}
billboard_long <- billboard_subset %>%
  pivot_longer(cols = starts_with("wk"),
               names_to = "week",
               names_prefix = "wk",
               values_to = "rank")
print(billboard_long)
```
...then back to wide
```{r pivot 3}
billboard_wide <- billboard_long %>%
  pivot_wider(names_from = "week",
              values_from = "rank",
              names_prefix = "wk")
print(billboard_wide)
```



# Managing missing values
Some data sets may have missing values, generally represented by `NA`. 
These can be due to:
- Missed medical appointment
- Error of measuring instruments
- Sampling stochasticity
- ...

```{r who2}
View(who2)
```
We can deal with these misisng values in multiple ways.

## 1- Ignoring missing values
We want to have the average number and standard deviation of cases per country for positive smear in males between 0 and 14 years old (sp_m_014)
```{r ignoring na 1}
who2 %>%
  group_by(country) %>%
  summarize(Mean = mean(sp_m_014),
            SD = sd(sp_m_014))
```
With na.rm
```{r ignoring na 2}
who2 %>%
  group_by(country) %>%
  summarize(Mean = mean(sp_m_014, na.rm = TRUE),
            SD = sd(sp_m_014, na.rm = TRUE))
```

## 2- Dropping missing values
We want to clean the table once and for all.
We can remove rows that have missing values in specified columns only
```{r dropping na 1}
who2 %>% drop_na(sp_m_014)
```
By default, drop_na will drop all rows with a missing value in any column
```{r dropping na 2}
who2 %>% drop_na()
```
Alternatively, we can drop rows that only have missing data
```{r dropping na 3}
to_keep <- who2[, c(-1, -2)] %>% # remove the first 2 columns
  is.na() %>% # test if each entry is an NA
  rowSums() != (ncol(who2)-2) # count the NAs in each row and check if different from the max
who2[to_keep,]
```

## 3- Replacing missing values
We assume NA replace the absence of cases and are equivalent to 0.
First, we create the named list used as parameter in replace_na()
```{r replacing na 1}
replacing_list <- rep(list(0), ncol(who2))
names(replacing_list) <- colnames(who2)
replacing_list['country'] <- 'unknown'
replacing_list['year'] <- 'unknown'
```
Then, we replace the NAs in the data frame
```{r replacing na 2}
who2 %>% replace_na(replacing_list)
```

## 4- Imputing missing values
We can replace missing values by custom values for each entry. Imputation consists in inferring missing data, generally using other datapoints as reference
Should be done on a case-by-case scenario, based on the data type and assumptions


## Hands-on 4: 
```{r hands on 4}
# 1- Drop the relapse columns
who2_ho4 <- who2[,grep('rel', colnames(who2), invert = TRUE)]
## 2- Remove empty rows 
## Optional: remove rows with at least 80% of the data missing
## 3- Pivot to long, then create new columns such that the data frame contains columns
## country, year, method, sex, age. Hint: search the doc!
## 4- Produce statistics of your choice
## 5- Re-pivot to wide
```
Solution
```{r hands on 4 ex}

```
