---
title: "Statistical testing"
author: 
- name: "Audrey Baguette"
  affiliation: Workshop Lead
- name: "Alex Diaz-Papkovich"
  affiliation: Original Material
- name: "Adrien Osakwe"
  affiliation: Original Material
date: "`r format(Sys.Date(), '%B %d, %Y')`"
---

# Loading libraries
```{r Loading libraries}
library(tidyverse)
```

# Two-samples t-test
1- First, we create two random samples from Normal distributions
```{r ttest 1}
set.seed(20221011) # Seed for reproducibility
n <- 100 # Sample size

mu1 <- 0 # Mean of sample 1
mu2 <- 0.1 # Mean of sample 2

sd1 <- 1 # Standard deviation 1
sd2 <- 1 # Standard deviation 2
```
Generate two samples of of size n of normally-distributed data.

Sample x1 has a mean of mu1 and standard deviation sd1

Sample x2 has a mean of mu2 and standard deviation sd2
```{r ttest 2}
x1 <- rnorm(n, mu1, sd1)
x2 <- rnorm(n, mu2, sd2)
```
2- We can take a look at the data with boxplots
```{r ttest 3}
x_df <- data.frame(Values = c(x1,x2),
                   Group = c(rep('G1',n),rep('G2',n)))
ggplot(x_df) + geom_boxplot(aes(y = Values, color = Group))
```
3- t-test
```{r ttest 4}
t.test(x1, x2)
```
4- The result of the test is influenced by the sample size
```{r ttest 5}
x3 <- rnorm(10000, mu1, sd1)
x4 <- rnorm(10000, mu2, sd2)
t.test(x3, x4)
```
4- We can also do a one-sided test
Check if group2 is greater than group1 (aka the difference 1-2 is smaller)
```{r ttest 6}
t.test(x1, x2, alternative = 'less')
```


# ANOVA
1- Create mutliple datasets

Our artificial data will have three different means
```{r anova 1}
set.seed(20221011)
m1 <- rnorm(30, mean = 25, sd = 5)
m2 <- rnorm(30, mean = 30, sd = 6)
m3 <- rnorm(30, mean = 22, sd = 7)

ms <- c(m1, m2, m3) # Combine our three measures into one vector

# Our vector of group labels A, B, and C
ls <- as.factor(c(rep("A",30),rep("B",30),rep("C",30)))
```
2- Visualize the data
```{r anova 2}
df_anova <- data.frame(Measure = ms, Group = ls)
ggplot(df_anova) + geom_boxplot(aes(y = Measure, color = Group))
```

3- ANOVA
```{r anova 3}
results_aov <- aov(Measure ~ Group, data = df_anova)
summary(results_aov)
```
We can look at the coefficients to find the difference across groups
```{r anova 4}
results_aov$coefficients
```


# Hands-on

## 1- ANOVA vs multiple t-tests
Data PlantGrowth has multiple groups: one control and two treatments

Make an ANOVA and consecutive t-tests between the control and each treatment. 
Do they lead to the same conclusion?
```{r hands on 1}
head(PlantGrowth)
```

Solution
```{r hands on 1 solution}
results_aov <- aov(weight ~ group, data = PlantGrowth)
summary(results_aov)
results_aov$coefficients
# At least one treatment is different, treatment 1 is lighter than control and
# treatment 2 is heavier

t.test(PlantGrowth$weight[PlantGrowth$group == 'ctrl'], PlantGrowth$weight[PlantGrowth$group == 'trt1'])
t.test(PlantGrowth$weight[PlantGrowth$group == 'ctrl'], PlantGrowth$weight[PlantGrowth$group == 'trt2'])
# Only treatment 2 is significantly different
```


## 2- Multiple testing correction
Take the co2 dataset (note: not CO2), that records the CO2 level for each month across several years. 
We want to test if the CO2 levels from 1959 to 1978 were lower than those between 1979 and 1997 for each month separately. 
Then, we correct for multiple testing and compare the results (hint: see p.adjust). 
Do the conclusions change? 
```{r hands on 2}
co2_df <- data.frame(matrix(co2, 39, 12), row.names = as.character(c(1959:1997)))
colnames(co2_df) <- c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec')
```

Solution
```{r hands on 2 solution}
p_values <- co2_df %>% apply(2, function(x) return(t.test(x[1:20], x[21:39])$p.value))
sum(p_values < 0.05) # 9 months are different
p_adj <- p.adjust(p_values, method = 'bonferroni')
sum(p_adj < 0.05) # 5 months are different
fdr <- p.adjust(p_values, method = 'fdr')
sum(fdr < 0.05) # 8 months are different
```

#